---
title: "01_Methanogen_prevalence_abundance"
author: "Mirabeau Mbong"
date: "2025-06-26"
output:
  html_notebook:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(dev = "pdf",
                      dpi = 300,
                      echo = FALSE,
                      cache = TRUE)

```

# load libraries

```{r}
# Load required libraries
library(phyloseq)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(vegan)
library(tibble)
options(bitmapType='cairo')
getwd()
```


```{r}
# Check working directory (should be repo root)
message("Current working directory: ", getwd())
```

# Functions: data processing I

```{r}
# ---------- Function to load Bracken + metadata and create phyloseq ----------
create_phyloseq <- function(bracken_file, metadata_file) {
  bracken <- read.table(bracken_file, header = TRUE, sep = "\t")
  meta <- read.table(metadata_file, header = TRUE, sep = "\t")

  # Process OTU table
  OTU <- bracken %>%
    dplyr::select(-c(taxonomy_id, taxonomy_lvl, taxonomy, taxIDs)) %>%
    dplyr::select(-contains('frac')) %>%
    dplyr::rename_with(~ sub("[_]num$", "", .x), ends_with("_num"))

  rownames(OTU) <- OTU$name
  OTU <- OTU %>% dplyr::select(-name)

  # Process taxonomy
  tax <- bracken %>%
    dplyr::select(name, taxonomy) %>%
    tidyr::separate(taxonomy,
                    into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
                    sep = ";", fill = "right", extra = "drop")
  rownames(tax) <- tax$name
  tax <- tax %>% dplyr::select(-name)

  # Process metadata
  meta$sample <- meta$ParticipantID
  rownames(meta) <- meta$sample
  meta <- meta %>% dplyr::select(-sample)

  # Create phyloseq object
  ps <- phyloseq(
    otu_table(as.matrix(OTU), taxa_are_rows = TRUE),
    tax_table(as.matrix(tax)),
    sample_data(meta)
  )
  return(ps)
}

# ---------- Usage of phyloseq function ----------

# ---------- Load data ----------
ps_adults_raw <- create_phyloseq("data/bracken_output_adults.tsv", "data/filtered_adult_metadata.txt")
ps_children_raw <- create_phyloseq("data/bracken_output_children.tsv", "data/filtered_child_metadata.txt")


# ---------- Function to filter and clean methanogens ----------

# Filter by phylum: Methanobacteriota
filter_methanogen_phylum <- function(ps) {
  subset_taxa(ps, Phylum == "p__Methanobacteriota")
}

# ---------- Clean and filter Methanobrevibacter_A genus ----------
filter_clean_methanobrevibacter_A <- function(ps) {
  ps_meth <- subset_taxa(ps, Genus == "g__Methanobrevibacter_A")
  
  tax_df <- as.data.frame(tax_table(ps_meth))
  tax_df$Species[is.na(tax_df$Species)] <- ""
  keep_taxa <- rownames(tax_df)[!grepl("sp", tax_df$Species)]
  ps_clean <- prune_taxa(keep_taxa, ps_meth)
  
  tax <- tax_table(ps_clean)
  tax[, "Genus"] <- gsub("^g__", "", tax[, "Genus"])
  tax[, "Species"] <- gsub("^s__", "", tax[, "Species"])
  tax[, "Species"] <- gsub("Methanobrevibacter_A smithii_A", 
                           "Methanobrevibacter_A intestini", tax[, "Species"])
  tax_table(ps_clean) <- tax
  
  return(ps_clean)
}

# ---------- usage of fitering function ----------

# Adults
methanogens1 <- filter_methanogen_phylum(ps_adults_raw)
methanobrevibacter1 <- filter_clean_methanobrevibacter_A(methanogens1)

# Children
methanogens2 <- filter_methanogen_phylum(ps_children_raw)
methanobrevibacter2 <- filter_clean_methanobrevibacter_A(methanogens2)

# ---------- All samples prevalence Computation ----------
compute_prevalence <- function(ps_obj, total_samples) {
  # Calculate prevalence per taxon
  prevalence_counts <- apply(
    X = otu_table(ps_obj),
    MARGIN = ifelse(taxa_are_rows(ps_obj), 1, 2),
    FUN = function(x) sum(x > 0)
  )
  
  # Build data.frame with prevalence, abundance, and taxonomy
  prev_df <- data.frame(
    Prevalence = prevalence_counts,
    TotalAbundance = taxa_sums(ps_obj),
    tax_table(ps_obj)
  )
  
  # Calculate percent prevalence
  prev_df <- prev_df %>%
    dplyr::mutate(prevalence = (Prevalence / total_samples) * 100)
  
  return(prev_df)
}

# ---------- Usage of prevalence function ----------

# For adults (510 samples)
prev_adults <- compute_prevalence(methanogens1, total_samples = 510)

# For children (394 samples)
prev_children <- compute_prevalence(methanogens2, total_samples = 394)

ps_adults_raw
ps_children_raw
```


# Functions: data processing II


```{r}
# ---------- compute species prevalence function ----------

compute_prevalence_by_group <- function(physeq_obj, group_var = "CurrCountry", abundance_threshold = 1e-1) {
  # Filter abundant taxa
  physeq_filtered <- filter_taxa(physeq_obj, function(x) mean(x) > abundance_threshold, TRUE)
  
  # Melt OTU table
  otu <- otu_table(physeq_filtered)
  if (!taxa_are_rows(physeq_filtered)) otu <- t(otu)
  otu_df <- as.data.frame(otu) %>% rownames_to_column("Taxon") %>%
    pivot_longer(-Taxon, names_to = "SampleID", values_to = "Abundance")
  
  # Add metadata
  sample_meta <- data.frame(sample_data(physeq_filtered)) %>% rownames_to_column("SampleID")
  otu_long <- left_join(otu_df, sample_meta, by = "SampleID")
  
  # Add taxonomy
  tax_df <- as.data.frame(tax_table(physeq_filtered)) %>% rownames_to_column("Taxon")
  otu_long <- left_join(otu_long, tax_df, by = "Taxon")
  
  # Compute prevalence
  prevalence_df <- otu_long %>%
    group_by(!!sym(group_var), Taxon, Genus, Species) %>%
    summarise(prevalence = mean(Abundance > 0) * 100, .groups = "drop") %>%
    mutate(
      Genus = str_remove(Genus, "^g__"),
      Species = str_remove(Species, "^s__"),
      Species = str_replace_all(Species, "Methanobrevibacter_A", "M.")
    )
  
  names(prevalence_df)[1] <- group_var  # Ensure group var is named properly
  return(prevalence_df)
}

# ---------- compute genus prevalence function --------

compute_genus_prevalence_by_group <- function(
  physeq_obj,
  group_var = "CurrCountry",
  abundance_threshold = 1e-1
) {

  # Filter abundant taxa
  physeq_filtered <- filter_taxa(
    physeq_obj,
    function(x) mean(x) > abundance_threshold,
    TRUE
  )

  # OTU table to long format
  otu <- otu_table(physeq_filtered)
  if (!taxa_are_rows(physeq_filtered)) otu <- t(otu)

  otu_df <- as.data.frame(otu) %>%
    tibble::rownames_to_column("Taxon") %>%
    tidyr::pivot_longer(
      -Taxon,
      names_to = "SampleID",
      values_to = "Abundance"
    )

  # Sample metadata
  sample_meta <- data.frame(sample_data(physeq_filtered)) %>%
    tibble::rownames_to_column("SampleID")

  otu_long <- dplyr::left_join(otu_df, sample_meta, by = "SampleID")

  # Taxonomy
  tax_df <- as.data.frame(tax_table(physeq_filtered)) %>%
    tibble::rownames_to_column("Taxon")

  otu_long <- dplyr::left_join(otu_long, tax_df, by = "Taxon")

  # Clean taxonomy labels
  otu_long <- otu_long %>%
    dplyr::mutate(
      Genus = stringr::str_remove(Genus, "^g__"),
      Species = stringr::str_remove(Species, "^s__"),
      Species = stringr::str_replace_all(Species, "Methanobrevibacter_A", "M.")
    )

  # ---- TRUE GENUS PRESENCE PER SAMPLE ----
  genus_sample_presence <- otu_long %>%
    dplyr::group_by(
      SampleID,
      !!rlang::sym(group_var),
      Genus
    ) %>%
    dplyr::summarise(
      present = any(Abundance > 0),
      .groups = "drop"
    )

  # ---- GENUS PREVALENCE PER GROUP ----
  prevalence_df <- genus_sample_presence %>%
    dplyr::group_by(
      !!rlang::sym(group_var),
      Genus
    ) %>%
    dplyr::summarise(
      prevalence = mean(present) * 100,
      .groups = "drop"
    )

  names(prevalence_df)[1] <- group_var
  return(prevalence_df)
}

# ---------- compute abundance function ----------

prepare_methanogen_abundance_df <- function(physeq_obj) {
  relabund_obj <- transform_sample_counts(physeq_obj, function(x) x / sum(x))
  df_long <- psmelt(relabund_obj)

  species_df <- df_long %>%
    filter(OTU %in% c("s__Methanobrevibacter_A smithii", "s__Methanobrevibacter_A smithii_A")) %>%
    mutate(
      Species = case_when(
        OTU == "s__Methanobrevibacter_A smithii" ~ "M. smithii",
        OTU == "s__Methanobrevibacter_A smithii_A" ~ "M. intestini"
      )
       # Pseudo-count for log scale if needed (Abundance = Abundance + 1e-7)
    )

  return(species_df)
}

# -------- compute abundances for phylum Phylum == "p__Methanobacteriota" -------

prepare_methanogen_abundance_phylum <- function(physeq_obj) {
  relabund_obj <- transform_sample_counts(physeq_obj, function(x) x / sum(x))
  df_long <- psmelt(relabund_obj)

  species_df <- df_long %>%
    filter(Phylum %in% c("p__Methanobacteriota")) %>%
    mutate(
      Species = case_when(
        OTU == "s__Methanobrevibacter_A smithii" ~ "M. smithii",
        OTU == "s__Methanobrevibacter_A smithii_A" ~ "M. intestini"
      )
       # Pseudo-count for log scale if needed (Abundance = Abundance + 1e-7)
    )

  return(species_df)
}

```


#  prevalence and abundance plotting functions


```{r}
# ---------- Custom color schemes ----------
country_colors <- c(Gabon = "#ff0000", Germany = "#e7e70e", Vietnam = "#0b0bd3")
species_colors <- c("Methanobrevibacter_A smithii" = "#1f77b4", "Methanobrevibacter_A intestini" = "#ff7f0e")
location_colors <- c("Lambaréné" = "skyblue", "Ikobey" = "orange")

plot_genus_prevalence <- function(prev_df) {
  prev_df_clean <- prev_df %>%
    filter(prevalence > 0) %>%  # Remove zero prevalence to avoid log10(0)
    mutate(
      Genus = gsub("^g__", "", Genus),
      Genus = gsub("^s__", "", Genus),
      Genus = ifelse(Genus == "Methanobrevibacter_A smithii_A", "Methanobrevibacter_A intestini", Genus)
    )

  # Compute maximum prevalence and relevel Genus by descending max prevalence
  max_prev_order <- prev_df_clean %>%
    group_by(Genus) %>%
    summarise(max_prev = max(prevalence, na.rm = TRUE)) %>%
    arrange(desc(max_prev)) %>%
    filter(!is.na(Genus))

  # Reorder factor levels BEFORE plotting
  prev_ordered <- prev_df_clean %>%
    filter(Genus %in% max_prev_order$Genus) %>%
    mutate(Genus = factor(Genus, levels = max_prev_order$Genus))

  # Plot
  ggplot(prev_ordered, aes(x = Genus, y = prevalence)) +
    geom_segment(aes(xend = Genus, y = 0.1, yend = prevalence), color = "grey60", linewidth = 0.7) +
    geom_point(color = "black", size = 2) +
    scale_y_log10(
      limits = c(0.1, 100),
      breaks = c(0.1, 1, 5, 10, 25, 50, 100),
      labels = c("0.1", "1", "5", "10", "25", "50", "100")
    ) +
    scale_x_discrete(labels = function(x) parse(text = paste0("italic('", x, "')"))) +
    labs(x = "", y = "Prevalence (log scale %)") +
    theme_light(base_size = 7, base_family = "sans") +
    theme(
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
      strip.text = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 7),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()
    )
}

plot_species_prevalence <- function(prev_df, genus_filter = "Methanobrevibacter_A", min_prevalence = 0.01) {
  prev_species <- prev_df %>%
    mutate(
      Genus = gsub("^g__", "", Genus),
      Species = gsub("^s__", "", Species),
      Species = gsub("^Methanobrevibacter_A smithii_A$", "Methanobrevibacter_A intestini", Species)
    ) %>%
    filter(
      Genus == genus_filter,
      !grepl("sp", Species),           # Remove MAGs
      prevalence > min_prevalence
    ) %>%
    mutate(
      Species = gsub(paste0("^", genus_filter), "M.", Species),
      Species = factor(Species, levels = rev(Species[order(prevalence)]))
    )
  
  ggplot(prev_species, aes(x = Species, y = prevalence)) +
    geom_segment(aes(xend = Species, y = 0, yend = prevalence), color = "grey60", linewidth = 0.7) +
    geom_point(color = "black", size = 2) +
    scale_y_continuous(
      limits = c(0, 100),
      breaks = seq(0, 100, 10),
      expand = c(0, 0),
      labels = function(x) x
    ) +
    scale_x_discrete(labels = function(x) parse(text = paste0("italic('", x, "')"))) +
    labs(x = "", y = "Prevalence (%)") +
    theme_light(base_size = 7, base_family = "sans") +
    theme(
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
      strip.text = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 7),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()
    )
}

# ---------- plot genus prevalence function ----------

plot_genus_prevalence_bar <- function(
  prev_df,
  group_var = "CurrCountry"
) {

  genus_df <- prev_df %>%
    dplyr::filter(!is.na(Genus), prevalence > 0) %>%
    dplyr::mutate(
      Genus = factor(
        Genus,
        levels = rev(unique(Genus[order(prevalence, decreasing = TRUE)]))
      )
    )

  p <- ggplot(genus_df,
              aes(x = Genus, y = prevalence, fill = !!rlang::sym(group_var))) +
    geom_bar(
      stat = "identity",
      position = position_dodge(width = 0.8),
      width = 0.6
    ) +
    coord_flip() +
    scale_y_sqrt(
      limits = c(0, 100),
      breaks = c(0, 1, 5, 10, 25, 50, 75, 100),
      labels = function(x) x
    ) +
    scale_fill_manual(
      name = "",
      values = country_colors
    ) +
    scale_x_discrete(
      labels = function(x) parse(text = paste0("italic('", x, "')"))
    ) +
    labs(
      x = "",
      y = "Genus prevalence (%)"
    ) +
    theme_light(base_size = 7, base_family = "sans") +
    theme(
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.position = "none"
    )

  return(p)
}

# ---------- plot species prevalence function ----------

plot_methanobrevibacter_species_bar <- function(prev_df, group_var = "CurrCountry", min_prevalence = 0.01) {
  meth_df <- prev_df %>%
    filter(Genus == "Methanobrevibacter_A", !str_detect(Species, "sp[0-9]+"), prevalence > min_prevalence) %>%
    mutate(
      Species = str_replace(Species, "M\\. smithii_A", "M. intestini")
    )
  
  if (nrow(meth_df) == 0) {
    warning("No eligible Methanobrevibacter species found.")
    return(NULL)
  }

  species_levels <- meth_df %>%
    group_by(Species) %>%
    summarise(median_prev = median(prevalence, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(median_prev)) %>%
    pull(Species)
  
  meth_df$Species <- factor(meth_df$Species, levels = rev(species_levels))

  ymax <- ceiling(max(meth_df$prevalence, na.rm = TRUE) / 10) * 10

  p <- ggplot(meth_df, aes(x = Species, y = prevalence, fill = !!sym(group_var))) +
    geom_bar(stat = "identity", position = position_dodge2(width = 0.7, preserve = "single"), width = 0.6) +
    coord_flip() +
    scale_y_continuous(
      limits = c(0, 100),
      breaks = seq(0, 100, 10),
      labels = function(x) x  # <- numeric labels only
    ) +
    scale_fill_manual(name = "", values = country_colors) +
    scale_x_discrete(labels = function(x) parse(text = paste0("italic('", x, "')"))) +
    labs(x = "", y = "Prevalence (%)", fill = group_var) +  # <- Updated y-axis label
    theme_light(base_size = 7, base_family = "sans") +
    theme(
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
      strip.text = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 7),
      legend.position = "none"
    )

  return(p)
}

# --------- ----------  #

```

# generate and save all prevalence data (Extended Data Fig. 2C & D)

```{r}
# Adults
plg_adults = plot_genus_prevalence(prev_adults)
pls_adults = plot_species_prevalence(prev_adults)

# Children
plg_children = plot_genus_prevalence(prev_children)
pls_children = plot_species_prevalence(prev_children)

# Save as PNG
ggsave("Extended-data-Fig2_species_prevalence_adults.png", plot = pls_adults, width = 6, height = 4, dpi = 300)

# Save as PDF
ggsave("Extended-data-Fig2_species_prevalence_adults.pdf", plot = pls_adults, width = 6, height = 4, dpi = 300)


ggsave("Extended-data-Fig2_species_prevalence_children.pdf", pls_children, width = 90, height = 95, units = "mm", dpi = 300)
ggsave("Extended-data-Fig2_species_prevalence_children.svg", pls_children, width = 90, height = 95, units = "mm")

plg_adults
pls_adults
plg_children
pls_children
```

# generate and save country specific prevalence (Extended Data Fig. 2A & B)


```{r}
# Species prevalence by country (adults):
prev_by_country <- compute_prevalence_by_group(methanogens1, group_var = "CurrCountry")
p_species_country <- plot_methanobrevibacter_species_bar(prev_by_country, group_var = "CurrCountry")

# Genus prevalence by country (adults):
prev_by_country_a <- compute_genus_prevalence_by_group(methanogens1, group_var = "CurrCountry")
p_genus_country <- plot_genus_prevalence_bar(prev_by_country_a, group_var = "CurrCountry")


# Species prevalence by country (children):
prev_by_country1 <- compute_prevalence_by_group(methanogens2, group_var = "CurrCountry")
p_species_country1 <- plot_methanobrevibacter_species_bar(prev_by_country1, group_var = "CurrCountry")

# Genus prevalence by country (children):
prev_by_country_c <- compute_genus_prevalence_by_group(methanogens2, group_var = "CurrCountry")
p_genus_country1 <- plot_genus_prevalence_bar(prev_by_country_c, group_var = "CurrCountry")


# Save as PNG
ggsave("Extended-data-Fig2_genus_prevalence_adults-by-country.png", plot = p_genus_country, width = 6, height = 4, dpi = 300)

# Save as PDF
ggsave("Extended-data-Fig2_genus_prevalence_adults-by-country.pdf", plot = p_genus_country, width = 6, height = 4, dpi = 300)

ggsave("Extended-data-Fig2_species_prevalence_children-by-country-2.pdf", p_species_country1, width = 140, height = 95, units = "mm", dpi = 300)
ggsave("Extended-data-Fig2_species_prevalence_children-by-country-2.svg", p_species_country1, width = 140, height = 95, units = "mm", dpi = 300, device = svglite::svglite)


p_genus_country
p_species_country
p_genus_country1
p_species_country1

```

# Get summary tables

```{r}
summary_table1 <- prev_by_country %>%
  filter(Species %in% c("M. smithii", "M. smithii_A")) %>%
  select(CurrCountry, Species, prevalence)

print(summary_table1)

summary_table2 <- prev_by_country1 %>%
  filter(Species %in% c("M. smithii", "M. smithii_A")) %>%
  select(CurrCountry, Species, prevalence)

print(summary_table2)

methanogens1
methanogens2

summary_table3 <- prev_adults %>%
  filter(Species %in% c("s__Methanobrevibacter_A smithii", "s__Methanobrevibacter_A smithii_A")) %>%
  select(Species, prevalence)

summary_table4 <- prev_children %>%
  filter(Species %in% c("s__Methanobrevibacter_A smithii", "s__Methanobrevibacter_A smithii_A")) %>%
  select(Species, prevalence)

summary_table3
summary_table4

# Add sampleType to each table
prev_by_country <- prev_by_country %>%
  mutate(sampleType = "Adult")

prev_by_country1 <- prev_by_country1 %>%
  mutate(sampleType = "Infant")

# Combine both tables
combined_prev <- bind_rows(prev_by_country, prev_by_country1) %>%
  # Select and reorder columns
  select(Country = CurrCountry, Species, prevalence, sampleType) %>%
  filter(prevalence > 0) 

combined_prev

# Write to a tab-delimited text file
write.table(combined_prev, file = "combined_prevalence_methanogens.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

# Genus prevalence by country (Extended Data Fig. 2A)

```{r}
write_genus_max_prevalence <- function(
  prev_df,
  group_var = "CurrCountry",
  out_file = "genus_max_prevalence_per_population.tsv"
) {

  # Global max prevalence per genus
  genus_global_max <- prev_df %>%
    dplyr::group_by(Genus) %>%
    dplyr::summarise(
      max_prevalence = max(prevalence),
      .groups = "drop"
    )

  # Country-specific max prevalence 
  genus_country_wide <- prev_df %>%
    dplyr::mutate(
      country_col = dplyr::case_when(
        !!rlang::sym(group_var) == "Gabon"    ~ "max_prevalence_Gabon",
        !!rlang::sym(group_var) == "Germany"  ~ "max_prevalence_Germany",
        !!rlang::sym(group_var) == "Vietnam"  ~ "max_prevalence_Vietnam",
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::filter(!is.na(country_col)) %>%
    dplyr::select(Genus, country_col, prevalence) %>%
    tidyr::pivot_wider(
      names_from  = country_col,
      values_from = prevalence
    )

  final_df <- genus_global_max %>%
    dplyr::left_join(genus_country_wide, by = "Genus") %>%
    dplyr::arrange(dplyr::desc(max_prevalence))

  readr::write_tsv(final_df, out_file)
  return(final_df)
}

genus_max_prev <- write_genus_max_prevalence(
  prev_df = prev_by_country_a,
  group_var = "CurrCountry",
  out_file = "genus_max_prevalence_per_population_adults.tsv"
)

genus_max_prev1 <- write_genus_max_prevalence(
  prev_df = prev_by_country_c,
  group_var = "CurrCountry",
  out_file = "genus_max_prevalence_per_population_children.tsv"
)

genus_max_prev_combined <- bind_rows(
  genus_max_prev  %>% mutate(sampleType = "adult"),
  genus_max_prev1 %>% mutate(sampleType = "infant")
)

genus_max_prev_combined <- genus_max_prev_combined %>%
  mutate(across(where(is.numeric), ~ tidyr::replace_na(.x, 0))) %>%
  select(sampleType, Genus, max_prevalence, max_prevalence_Gabon, max_prevalence_Germany, max_prevalence_Vietnam)

readr::write_tsv(
  genus_max_prev_combined,
  "genus_max_prev_combined.txt"
)

genus_max_prev_combined

#colnames(genus_max_prev_combined)
```

### Species prevalence by country

```{r}
compute_prevalence_counts_by_group <- function(physeq_obj, group_var = "CurrCountry", abundance_threshold = 1e-1) {
  # Filter taxa by average abundance
  physeq_filtered <- filter_taxa(physeq_obj, function(x) mean(x) > abundance_threshold, TRUE)

  # Extract OTU table
  otu_mat <- otu_table(physeq_filtered)
  if (!taxa_are_rows(physeq_filtered)) otu_mat <- t(otu_mat)

  # Create long table of sample-taxon abundances
  otu_long <- as.data.frame(otu_mat) %>%
    rownames_to_column("Taxon") %>%
    pivot_longer(-Taxon, names_to = "SampleID", values_to = "Abundance")

  # Sample metadata 
  sample_meta <- as(sample_data(physeq_filtered), "data.frame") %>%
    tibble::rownames_to_column("SampleID") %>%
    select(SampleID, all_of(group_var))

  # Taxonomy
  tax_df <- tax_table(physeq_filtered) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Taxon") %>%
    select(Taxon, Genus, Species)

  # Merge and clean taxonomy
  df <- otu_long %>%
    left_join(sample_meta, by = "SampleID") %>%
    left_join(tax_df, by = "Taxon") %>%
    mutate(
      Genus = str_remove(Genus, "^g__"),
      Species = str_remove(Species, "^s__"),
      Species = str_replace_all(Species, "Methanobrevibacter_A", "M.")
    ) %>%
    filter(!is.na(.data[[group_var]]))   

  # Summarize presence/absence per sample
  df_pa <- df %>%
    group_by(SampleID, .data[[group_var]], Species) %>%
    summarise(present = any(Abundance > 0), .groups = "drop")

  # Aggregate counts per group & species
  summary_counts <- df_pa %>%
    group_by(.data[[group_var]], Species) %>%
    summarise(
      n_positive = sum(present),
      n_total = n(),  
      n_negative = n_total - n_positive,
      prevalence = n_positive / n_total * 100,
      .groups = "drop"
    )

  # Rename column
  names(summary_counts)[1] <- group_var

  return(summary_counts)
}

```

```{r}
library(data.table)
# For adult methanogen phyloseq object
summary_with_counts_adults <- compute_prevalence_counts_by_group(methanogens1, group_var = "CurrCountry")

# For child dataset if you have one
summary_with_counts_children <- compute_prevalence_counts_by_group(methanogens2, group_var = "CurrCountry")

summary_with_counts_adults = as.data.table(summary_with_counts_adults)
summary_with_counts_children = as.data.table(summary_with_counts_children)

print(summary_with_counts_adults, nrow=81)
print(summary_with_counts_children, nrow=201)

write.table(summary_with_counts_adults,file = "/data/summary_with_counts_adults.txt", row.names = FALSE, sep = "\t")


```

### Chisquare country comparisons of species prevalence

```{r}
run_chisq_by_species <- function(summary_counts_df, group_var = "CurrCountry", p_adjust_method = "BH") {
  library(dplyr)
  library(rlang)
  group_var_sym <- sym(group_var)
  
  # Split by species
  species_list <- split(summary_counts_df, summary_counts_df$Species)
  
  chisq_results <- lapply(names(species_list), function(species_name) {
    df <- species_list[[species_name]]
    
    contingency <- df %>%
      select(!!group_var_sym, n_positive, n_negative) %>%
      column_to_rownames(var = group_var) %>%
      as.matrix()
    
    if (nrow(contingency) > 1 && all(rowSums(contingency) > 0)) {
      test <- suppressWarnings(chisq.test(contingency))
      tibble(
        Species = species_name,
        p_value = test$p.value,
        chi_squared = as.numeric(test$statistic),
        df = as.numeric(test$parameter)
      )
    } else {
      tibble(Species = species_name, p_value = NA, chi_squared = NA, df = NA)
    }
  })
  
  # Combine and adjust p-values
  results <- bind_rows(chisq_results) %>%
    mutate(
      p_adj = p.adjust(p_value, method = p_adjust_method)
    )
  
  return(results)
}
```

```{r}
# Filter for the two species of interest
filtered_summary1 <- summary_with_counts_adults %>%
  filter(Species %in% c("M. smithii", "M. smithii_A"))

# Filter for the two species of interest
filtered_summary2 <- summary_with_counts_children %>%
  filter(Species %in% c("M. smithii", "M. smithii_A"))

# For adults
chisq_results_adults <- run_chisq_by_species(filtered_summary1, group_var = "CurrCountry", p_adjust_method = "BH")

# For children (if available)
chisq_results_children <- run_chisq_by_species(filtered_summary2, group_var = "CurrCountry", p_adjust_method = "BH")

chisq_results_adults
chisq_results_children

```


# obtain abundances M. smithii/M. intestini by country & sampling locations

```{r}
# Prepare data
adults_df <- prepare_methanogen_abundance_df(ps_adults_raw)
children_df <- prepare_methanogen_abundance_df(ps_children_raw)

```

## obtain abundances of methanogens (Phylum Methanobacteriota) by country 


```{r}
adults_met_df <- prepare_methanogen_abundance_phylum(ps_adults_raw)
children_met_df <- prepare_methanogen_abundance_phylum(ps_children_raw)


adults_met_df1 <- adults_met_df %>%
  select(-Species) %>%
  mutate(
    Species = case_when(
      str_detect(OTU, "^s__Methanobrevibacter_A smithii_A") ~ "M. intestini",

      # Other Methanobrevibacter_A entries
      str_detect(OTU, "^s__Methanobrevibacter_A") ~
        str_replace(OTU, "^s__Methanobrevibacter_A\\s*", "M. "),

      # All other species: just remove s__
      TRUE ~ str_remove(OTU, "^s__")
    )
  )

children_met_df1 <- children_met_df %>%
  select(-Species) %>%
  mutate(
    Species = case_when(
      str_detect(OTU, "^s__Methanobrevibacter_A smithii_A") ~ "M. intestini",

      # Other Methanobrevibacter_A entries
      str_detect(OTU, "^s__Methanobrevibacter_A") ~
        str_replace(OTU, "^s__Methanobrevibacter_A\\s*", "M. "),

      # All other species: remove s__
      TRUE ~ str_remove(OTU, "^s__")
    )
  )

adults_met_df1
children_met_df1

summarize_species_abundance <- function(df, 
                                        sample_col = "Sample",
                                        country_col = "CurrCountry",
                                        species_col = "Species",
                                        abundance_col = "Abundance") {
  
  df_summary <- df %>%
    group_by(!!sym(country_col), !!sym(species_col)) %>%
    summarise(
      mean_abundance = mean(!!sym(abundance_col), na.rm = TRUE),
      median_abundance = median(!!sym(abundance_col), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    # Rename for clarity
    rename(Country = !!sym(country_col), Species = !!sym(species_col))
  
  return(df_summary)
}

mean_abund_adults = summarize_species_abundance(adults_met_df1)
mean_abund_child = summarize_species_abundance(children_met_df1)

mean_abund_adults
mean_abund_child
prev_by_country 


prev_by_country <- prev_by_country %>%
  mutate(sampleType = "Adult") %>%
  # Select and reorder columns
  select(Country = CurrCountry, Species, prevalence, sampleType) %>%
  mutate(
    Species = if_else(
      Species == "M. smithii_A",
      "M. intestini",
      Species
    )
  )

prev_by_country1 <- prev_by_country1 %>%
  mutate(sampleType = "Infant") %>%
  # Select and reorder columns
  select(Country = CurrCountry, Species, prevalence, sampleType) %>%
  mutate(
    Species = if_else(
      Species == "M. smithii_A",
      "M. intestini",
      Species
    )
  )

#join tables
# adults
prev_by_country_enriched_adults <- prev_by_country %>%
  left_join(
    mean_abund_adults,
    by = c("Country", "Species")
  )
# infants
prev_by_country_enriched_infants <- prev_by_country1 %>%
  left_join(
    mean_abund_child,
    by = c("Country", "Species")
  )

# Combine both tables
combined_prev_abund <- bind_rows(prev_by_country_enriched_adults, prev_by_country_enriched_infants) %>%
  # Select and reorder columns
  select(Country, sampleType, Species, prevalence, mean_abundance, median_abundance) #%>%
  #filter(prevalence > 0)  # keep only rows where prevalence is greater than 0

combined_prev_abund

# Write to a tab-delimited text file
write.table(combined_prev_abund, file = "/data/combined_prevalence_abundance_methanogens.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```


#  function to plot methanogen abundance

```{r}
plot_methanogen_abundance <- function(
  species_df,
  group_var,
  comparisons_list,
  xlab = "",
  country_colors = NULL
) {
  library(ggplot2)
  library(ggpubr)
  library(dplyr)

  # Add a small pseudocount to allow log10 scaling without losing zeros
  # (keeps zeros visible without distorting higher abundances)
  df_plot <- species_df %>%
    mutate(Abundance_plot = Abundance + 1e-7)  # adjust if necessary

  # Determine x-axis label
  x_label <- if (is.null(xlab) || xlab == "") group_var else xlab

  # Split by species and plot
  plots <- df_plot %>%
    split(.$Species) %>%
    lapply(function(df_species) {
      species_name <- unique(df_species$Species)

      ggplot(df_species, aes_string(x = group_var, y = "Abundance_plot * 100", color = group_var)) +
        geom_boxplot(outlier.shape = NA, fill = NA, color = "black", alpha = 0.8) +
        geom_jitter(width = 0.2, alpha = 0.6, size = 1) +
        #stat_summary(
          #fun = median, geom = "text",
          #aes(label = round(..y.., 3)),
          #color = "blue", vjust = -0.5, size = 2.5
        #) +
        stat_compare_means(
          label = "p.signif", method = "wilcox.test",
          comparisons = comparisons_list, tip.length = 0, vjust = 0.3
        ) +
        scale_y_log10(labels = scales::label_number(accuracy = 0.0001)) +
        scale_color_manual(values = country_colors, na.translate = FALSE) +
        labs(
          y = "Relative abundance (log10 %)",
          x = NULL,
          title = species_name
        ) +
        theme_light(base_size = 7, base_family = "sans") +
        theme(
          plot.title = element_text(size = 8, hjust = 0.5, face = "italic"),
          axis.text = element_text(size = 7),
          axis.title = element_text(size = 7),
          legend.position = "none"
        )
    })

  return(plots)
}

```


```{r}
# Your color palette (e.g. for countries)
country_colors <- c("Gabon" = "#ff0000", "Germany" = "#e7e70e", "Vietnam" = "#0b0bd3")

# Example comparisons (adjust as needed)
comparisons_list <- list(c("Gabon", "Germany"), c("Gabon", "Vietnam"), c("Germany", "Vietnam"))

# Call the function
species_plots_adults <- plot_methanogen_abundance(adults_df, group_var = "CurrCountry",
                                         comparisons_list = comparisons_list,
                                         xlab = "", country_colors = country_colors)


# Show individual plots
species_plots_adults[["M. smithii"]]
species_plots_adults[["M. intestini"]]

ggsave("M_intestini_abundance_adults-Fig1.pdf", species_plots_adults[["M. intestini"]], width = 80, height = 70, units = "mm", device = cairo_pdf, dpi = 300)
ggsave("M_intestini_abundance_adults-Fig1.svg", species_plots_adults[["M. intestini"]], width = 80, height = 70, units = "mm", dpi = 300, device = svglite::svglite)

```


```{r}
species_plots_children <- plot_methanogen_abundance(children_df, group_var = "CurrCountry",
                                         comparisons_list = comparisons_list,
                                         xlab = "", country_colors = country_colors)

# Show individual plots
species_plots_children[["M. smithii"]]
species_plots_children[["M. intestini"]]

ggsave("M_intestini_abundance_children-Fig1.pdf", species_plots_children[["M. intestini"]], width = 50, height = 60, units = "mm", device = cairo_pdf, dpi = 300)
ggsave("M_smithii_abundance_children-Fig1.svg", species_plots_children[["M. smithii"]], width = 80, height = 70, units = "mm", dpi = 300, device = svglite::svglite)

```


### Extract country medians and comparison stats

```{r}
summarise_abundance_stats <- function(df, group_var = "CurrCountry") {
  # Compute median abundance per group and species
  medians <- df %>%
    group_by(Species, !!sym(group_var)) %>%
    summarise(median_abundance = median(Abundance) * 100, .groups = "drop")
  
  # Compute pairwise Wilcoxon tests per species
  pvals <- df %>%
    group_split(Species) %>%
    map_df(function(species_df) {
      species_name <- unique(species_df$Species)
      
      test_res <- pairwise.wilcox.test(
        x = species_df$Abundance,
        g = species_df[[group_var]],
        p.adjust.method = "BH"
      )
      
      # Convert matrix to tidy data frame
      pval_matrix <- as.data.frame(as.table(test_res$p.value))
      colnames(pval_matrix) <- c("group1", "group2", "p_adj")
      
      # Add species label
      pval_matrix$Species <- species_name
      
      # Remove NAs
      pval_matrix %>%
        filter(!is.na(p_adj))
    })
  
  list(medians = medians, pvals = pvals)
}

```

# generate tables of medians and country comparison stats

```{r}
# Adults
adult_stats <- summarise_abundance_stats(adults_df)
medians_adults <- adult_stats$medians
pvals_adults <- adult_stats$pvals

# Children
child_stats <- summarise_abundance_stats(children_df)
medians_children <- child_stats$medians
pvals_children <- child_stats$pvals

# Median abundances
medians_adults %>% filter(Species == "M. smithii")
medians_children %>% filter(Species == "M. smithii")

medians_adults %>% filter(Species == "M. intestini")
medians_children %>% filter(Species == "M. intestini")

# P-values for Gabon vs Germany
#pvals_adults %>% filter(Species == "M. smithii", group1 == "Gabon", group2 == "Germany")

pvals_adults
pvals_children

```

# Co-occurence stats of M. smithii and M. intestini


```{r}
summarize_methanogen_cooccurrence_stats <- function(physeq_obj) {
  # Extract OTU table 
  otu <- as.data.frame(otu_table(physeq_obj))
  if (taxa_are_rows(physeq_obj)) otu <- t(otu)
  
  # Rename target species (GTDB names to friendly names)
  colnames(otu) <- case_when(
    colnames(otu) == "s__Methanobrevibacter_A smithii" ~ "M. smithii",
    colnames(otu) == "s__Methanobrevibacter_A smithii_A" ~ "M. intestini",
    TRUE ~ colnames(otu)
  )
  
  # Keep only methanogens of interest
  target_otus <- c("M. smithii", "M. intestini")
  otu_filtered <- otu[, intersect(target_otus, colnames(otu)), drop = FALSE]
  
  # Convert to binary presence/absence
  otu_pa <- as.data.frame((otu_filtered > 0) * 1)
  
  # Add metadata (country)
  sample_meta <- as.data.frame(sample_data(physeq_obj))
  otu_pa$CurrCountry <- sample_meta$CurrCountry
  
  # ---- 1. Summary counts + proportions ----
  summary_tbl <- otu_pa %>%
    mutate(pattern = case_when(
      `M. smithii` == 1 & `M. intestini` == 1 ~ "both",
      `M. smithii` == 1 & `M. intestini` == 0 ~ "M. smithii only",
      `M. smithii` == 0 & `M. intestini` == 1 ~ "M. intestini only",
      TRUE ~ "none"
    )) %>%
    group_by(CurrCountry, pattern) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(CurrCountry) %>%
    mutate(prop = round(100 * n / sum(n), 2)) %>%
    arrange(CurrCountry, pattern)
  
  # ---- 2. Chi-square tests per country ----
  chisq_results <- otu_pa %>%
    group_by(CurrCountry) %>%
    group_modify(~{
      tab <- table(.x$`M. smithii`, .x$`M. intestini`)
      # Ensure 2x2 before running
      if (all(dim(tab) == c(2, 2))) {
        res <- chisq.test(tab)
        tibble(
          statistic = res$statistic,
          df = res$parameter,
          p.value = res$p.value
        )
      } else {
        tibble(statistic = NA, df = NA, p.value = NA)
      }
    }) %>%
    ungroup()
  
  return(list(summary = summary_tbl, chisq = chisq_results))
}
```

# Summary stats (Co-occurence M. smithii/M. intestini)

```{r}
summary_adults <- summarize_methanogen_cooccurrence_stats(methanogens1)
summary_children <- summarize_methanogen_cooccurrence_stats(methanogens2)
summary_adults$summary
summary_adults$chisq
summary_children$summary
summary_children$chisq

```



# Abdundance plots by Sampling locations (Gabon Only)

```{r}
# Gabon only: filter and plot by sampling location
adults_gabon <- adults_df %>% filter(CurrCountry == "Gabon")
children_gabon <- children_df %>% filter(CurrCountry == "Gabon")
gabon_comparisons <- list(c("Lambarene", "Ikobey"))
location_colors <- c("Lambarene" = "skyblue", "Ikobey" = "orange")

species_plots_adults_loc <- plot_methanogen_abundance(
  adults_gabon,
  group_var = "samplingLocation",
  comparisons_list = gabon_comparisons,
  xlab = "",
  country_colors = location_colors
)

# Show individual plots
species_plots_adults_loc[["M. smithii"]]
species_plots_adults_loc[["M. intestini"]]

ggsave("M_intestini_abundance_adults-Gabon-Fig1.pdf", species_plots_adults_loc[["M. intestini"]], width = 50, height = 60, units = "mm", device = cairo_pdf, dpi = 300)
ggsave("M_smithii_abundance_adults-Gabon-Fig1.svg", species_plots_adults_loc[["M. smithii"]], width = 80, height = 70, units = "mm", dpi = 300, device = svglite::svglite)
```


```{r}
species_plots_children_loc <- plot_methanogen_abundance(
  children_gabon,
  group_var = "samplingLocation",
  comparisons_list = gabon_comparisons,
  xlab = "",
  country_colors = location_colors
)

# Show individual plots
species_plots_children_loc[["M. smithii"]]
species_plots_children_loc[["M. intestini"]]

ggsave("M_smithii_abundance_children-Gabon-Fig1.pdf", species_plots_children_loc[["M. smithii"]], width = 50, height = 60, units = "mm", dpi = 300)
ggsave("M_intestini_abundance_children-Gabon-Fig1.svg", species_plots_children_loc[["M. intestini"]], width = 80, height = 70, units = "mm", dpi = 300, device = svglite::svglite)
```

# Summary stats/median abundance M. smithii/M. intestini Gabon

```{r}

adult_gab_stats <- summarise_abundance_stats(adults_gabon, group_var = "samplingLocation")
medians_adults_gab <- adult_gab_stats$medians
pvals_adults_gab <- adult_gab_stats$pvals

children_gab_stats <- summarise_abundance_stats(children_gabon, group_var = "samplingLocation")
medians_children_gab <- children_gab_stats$medians
pvals_children_gab <- children_gab_stats$pvals

pvals_adults_gab
pvals_children_gab

medians_adults_gab
medians_children_gab
```


# Summary stats/median abundance M. smithii/M. intestini Vietnam


```{r}
adults_viet <- adults_df %>% filter(CurrCountry == "Vietnam")
children_viet <- children_df %>% filter(CurrCountry == "Vietnam")

adults_viet %>%
  distinct(Sample, samplingLocation) %>%   # ensures each sample counted once
  count(samplingLocation, name = "n_samples")
children_viet %>%
  distinct(Sample, samplingLocation) %>%   # ensures each sample counted once
  count(samplingLocation, name = "n_samples")


# BirthState, CurrState.Province

adult_viet_stats <- summarise_abundance_stats(adults_viet, group_var = "samplingLocation")
medians_adults_viet <- adult_viet_stats$medians
pvals_adults_viet <- adult_viet_stats$pvals

children_viet_stats <- summarise_abundance_stats(children_viet, group_var = "samplingLocation")
medians_children_viet <- children_viet_stats$medians
pvals_children_viet <- children_viet_stats$pvals

pvals_adults_viet
pvals_children_viet

medians_adults_viet
medians_children_viet

```


# Adults Vs Children M.smithii/M. intestini abundance plots

```{r}

# Prepare adult and child subsets
min_abundance_adults <- 0
adults_df_subset <- adults_df %>%
  #filter(Abundance > min_abundance_adults) %>%
  select(Sample, Abundance, samplingLocation, CurrCountry, Species) %>%
  mutate(Dataset = "Adults")

children_df_subset <- children_df %>%
  select(Sample, Abundance, samplingLocation, CurrCountry, Species) %>%
  mutate(Dataset = "Children")

# Combine
combined_df <- bind_rows(adults_df_subset, children_df_subset) %>%
  mutate(
    CountryGroup = paste(CurrCountry, Dataset),
    LocationGroup = paste(samplingLocation, Dataset)
  )
```


#  function to generate M. smithii/M. intestini abundance plots (Adults vs. Children)


```{r}
plot_methanogen_abundance_grouped <- function(df, group_var, comparisons_list, color_by = NULL) {
  # Add ColorGroup column based on group_var
  df <- df %>%
    mutate(ColorGroup = dplyr::case_when(
      group_var == "CountryGroup" ~ CurrCountry,
      group_var == "LocationGroup" ~ samplingLocation,
      TRUE ~ NA_character_
    ))

  # Define color map
  if (color_by == "country") {
    color_map <- country_colors
  } else if (color_by == "location") {
    color_map <- location_colors
  } else {
    color_map <- NULL
  }

  # Custom labeller for italic species names
  species_labeller <- function(labels) {
    sapply(labels, function(x) paste0("italic('", x, "')"))
  }

  species_list <- unique(df$Species)
  plot_list <- list()

  for (species_name in species_list) {
    df_species <- df %>% filter(Species == species_name)

    y.max <- max(df_species$Abundance * 100, na.rm = TRUE)
    y.position <- seq(y.max * 1.1, length.out = length(comparisons_list), by = y.max * 0.1)

    p <- ggplot(df_species, aes_string(x = group_var, y = "Abundance * 100")) +
      geom_boxplot(outlier.shape = NA, fill = NA, color = "black", alpha = 0.8) +
      geom_jitter(aes(color = ColorGroup), width = 0.3, alpha = 0.3) +
      scale_y_log10(labels = scales::label_number(accuracy = 0.0001)) +
      stat_compare_means(
        label = "p.signif",
        method = "wilcox.test",
        comparisons = comparisons_list,
        tip.length = 0,
        vjust = 0.3
      ) +
      scale_color_manual(values = color_map) +
      labs(
        y = "Relative abundance (log10 %)",
        x = NULL  
      ) +
      theme_light(base_size = 7, base_family = "sans") +
      theme(
        strip.text = element_text(size = 7),
        axis.text = element_text(size = 7),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.title = element_text(size = 7),
        text = element_text(size = 7, colour = "black"),
        legend.position = "none"
      )

    plot_list[[species_name]] <- p
  }

  return(plot_list)
}

```

# Generate and save plots


```{r}
country_comparisons <- list(
  c("Gabon Adults", "Gabon Children"),
  c("Germany Adults", "Germany Children"),
  c("Vietnam Adults", "Vietnam Children")
)

gabon_location_comparisons <- list(
  c("Lambarene Adults", "Lambarene Children"),
  c("Ikobey Adults", "Ikobey Children")
)

plot_country_age_list <- plot_methanogen_abundance_grouped(
  df = combined_df,
  group_var = "CountryGroup",
  comparisons_list = country_comparisons,
  color_by = "country"
)

# Location plots for Gabon
# Subset Gabon samples if needed
gabon_df <- combined_df %>% filter(CurrCountry == "Gabon")

plot_gabon_age_list <- plot_methanogen_abundance_grouped(
  df = gabon_df,
  group_var = "LocationGroup",
  comparisons_list = gabon_location_comparisons,
  color_by = "location"
)

plot_country_age_list
plot_gabon_age_list


# Save as PDF or PNG
#ggsave("smithii_age-group_by_country.pdf", plot_country_age_list[["M. smithii"]], width = 4, height = 4, dpi = 300)

ggsave("smithii_age-group_Gabon-extended-data-Fig2.pdf", plot_gabon_age_list[["M. smithii"]], width = 95, height = 85, units = "mm", dpi = 300)
ggsave("smithii_age-group_Gabon-extended-data-Fig2.svg", plot_gabon_age_list[["M. smithii"]], width = 95, height = 85, units = "mm", dpi = 300, device = svglite::svglite)
```

# summary stats/median abundance

```{r}
# for all countries
all_country_stats <- summarise_abundance_stats(combined_df, group_var = "Dataset")
medians_all_countries <- all_country_stats$medians
pvals_all_countries <- all_country_stats$pvals

### for gabon
gab_sp_stats <- summarise_abundance_stats(gabon_df, group_var = "Dataset")
medians_gab_sp <- gab_sp_stats$medians
pvals_gab_sp <- gab_sp_stats$pvals

medians_all_countries
pvals_all_countries

medians_gab_sp
pvals_gab_sp

```



# correlations between paired and unpaired mother-child

```{r}
pairing_df = read.table("/data/mother_child-sample-ids.txt", header = TRUE, sep = "\t")

```

# mother-child colonization function

```{r}

run_colonization_permutation_pa_binary <- function(
  combined_df,
  pairing_df,
  group_var = "CurrCountry",
  presence_threshold = 0,
  n_perm = 1000,
  min_pairs = 5,
  return_plots = FALSE,
  seed = 42
) {
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(ggplot2)

  set.seed(seed)

  # 1) Make presence/absence table
  combined_pa <- combined_df %>%
    mutate(
      Abundance = as.numeric(Abundance),
      Present = if_else(is.na(Abundance), 0L,
                        if_else(Abundance > presence_threshold, 1L, 0L))
    ) %>%
    select(Sample, Species, !!sym(group_var), Present)

  # 2) Mother–child pairs
  pairs_long <- pairing_df %>%
    pivot_longer(cols = starts_with("Infant"), names_to = "infant_col", values_to = "ChildID") %>%
    filter(!is.na(ChildID)) %>%
    select(MotherID = Sample, ChildID)

  valid_sample_ids <- unique(combined_pa$Sample)
  pairs_valid <- pairs_long %>%
    filter(MotherID %in% valid_sample_ids & ChildID %in% valid_sample_ids) %>%
    distinct()

  if (nrow(pairs_valid) == 0) {
    warning("No valid mother-child pairs found with matching samples in combined_df.")
    return(tibble(Species = character(), Country = character(), n_pairs = integer(),
                  observed_phi = numeric(), permuted_mean = numeric(), permuted_sd = numeric(),
                  p_permutation = numeric(), p_adj_BH = numeric()))
  }

  # 3) Join on species and sample to ensure 1:1 pairing
  pairs_species <- pairs_valid %>%
    crossing(Species = unique(combined_pa$Species)) %>% # expand to all species
    left_join(
      combined_pa %>%
        rename(Mother = Present, Mother_Country = !!sym(group_var)),
      by = c("MotherID" = "Sample", "Species")
    ) %>%
    left_join(
      combined_pa %>%
        rename(Child = Present, Child_Country = !!sym(group_var)),
      by = c("ChildID" = "Sample", "Species")
    ) %>%
    mutate(Country = coalesce(as.character(Mother_Country), as.character(Child_Country))) %>%
    select(Species, Country, MotherID, ChildID, Mother, Child) %>%
    filter(!is.na(Mother) & !is.na(Child)) %>%
    distinct()

  if (nrow(pairs_species) == 0) {
    warning("No mother-child pairs with matching species presence found.")
    return(tibble(Species = character(), Country = character(), n_pairs = integer(),
                  observed_phi = numeric(), permuted_mean = numeric(), permuted_sd = numeric(),
                  p_permutation = numeric(), p_adj_BH = numeric()))
  }
  

  # 4) Compute correlations
  species_country_groups <- pairs_species %>%
    group_by(Species, Country) %>%
    group_split()

  results_list <- list()
  plot_list <- list()

  for (grp in species_country_groups) {
    sp <- unique(grp$Species)
    ct <- unique(grp$Country)
    n_pairs_here <- nrow(grp)

    if (n_pairs_here < min_pairs) {
      results_list[[paste(sp, ct, sep = "_")]] <- tibble(
        Species = sp, Country = ct, n_pairs = n_pairs_here,
        observed_phi = NA_real_, permuted_mean = NA_real_, permuted_sd = NA_real_,
        p_permutation = NA_real_
      )
      next
    }

    mother_vec <- as.numeric(grp$Mother)
    child_vec  <- as.numeric(grp$Child)

    if (sd(mother_vec) == 0 || sd(child_vec) == 0) {
      results_list[[paste(sp, ct, sep = "_")]] <- tibble(
        Species = sp, Country = ct, n_pairs = n_pairs_here,
        observed_phi = NA_real_, permuted_mean = NA_real_, permuted_sd = NA_real_,
        p_permutation = NA_real_
      )
      next
    }

    observed_phi <- suppressWarnings(cor(mother_vec, child_vec, method = "pearson"))

    permuted_phis <- replicate(n_perm, {
      shuffled_child <- sample(child_vec)
      suppressWarnings(cor(mother_vec, shuffled_child, method = "pearson"))
    })

    perm_mean <- mean(permuted_phis, na.rm = TRUE)
    perm_sd <- sd(permuted_phis, na.rm = TRUE)
    p_perm <- mean(abs(permuted_phis) >= abs(observed_phi), na.rm = TRUE)

    results_list[[paste(sp, ct, sep = "_")]] <- tibble(
      Species = sp, Country = ct, n_pairs = n_pairs_here,
      observed_phi = observed_phi,
      permuted_mean = perm_mean,
      permuted_sd = perm_sd,
      p_permutation = p_perm
    )

    if (return_plots) {
      plt <- tibble(permuted_phis = permuted_phis) %>%
        ggplot(aes(x = permuted_phis)) +
        geom_histogram(bins = 30, fill = "gray80", color = "black") +
        geom_vline(xintercept = observed_phi, color = "red", linetype = "dashed", size = 1) +
        labs(
          title = paste(sp, "—", ct),
          subtitle = paste0("Observed phi = ", round(observed_phi, 3),
                            " | perm p = ", signif(p_perm, 3)),
          x = "Permuted phi (null)", y = "Frequency"
        ) +
        theme_minimal()
      plot_list[[paste(sp, ct, sep = "_")]] <- plt
    }
  }

  results_df <- bind_rows(results_list) %>%
    mutate(p_adj_BH = p.adjust(p_permutation, method = "BH"))

  if (return_plots) {
    return(list(stats = results_df, plots = plot_list, pairs_table = pairs_species))
  } else {
    return(list(stats = results_df, pairs_table = pairs_species))
  }
}

```

# generate stats

```{r}
res <- run_colonization_permutation_pa_binary(
  combined_df = combined_df,
  pairing_df = pairing_df,
  group_var = "CurrCountry",
  presence_threshold = 1e-7,   # any value ≤ this is "absent"
  n_perm = 1000,
  min_pairs = 5,
  return_plots = TRUE
)

# View results
head(res$stats)

# Example plot for one species-country
res$plots[["M. smithii_Gabon"]]

```



# assessing mutual exclusion M. smithii vs M. intestini

```{r}
# reshape from long to wide format
reshape_methanogen_abundances <- function(df_long) {
  df_long %>%
    filter(Species %in% c("M. smithii", "M. intestini")) %>%
    tidyr::pivot_wider(names_from = Species, values_from = Abundance) %>%
    rename(M_smithii = `M. smithii`, M_intestini = `M. intestini`) %>%
    mutate(
      M_smithii = replace_na(M_smithii, 0),
      M_intestini = replace_na(M_intestini, 0)
    )
}

# Density plot of log-ratio
plot_log_ratio_density <- function(df_wide, filename = "log_ratio_density.png") {
  df_wide <- df_wide %>%
    mutate(log_ratio = log10((M_smithii + 1e-5) / (M_intestini + 1e-5)))

  p <- ggplot(df_wide, aes(x = log_ratio)) +
    geom_density(fill = "skyblue", alpha = 0.6) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
      title = "Distribution of log₁₀(M. smithii / M. intestini)",
      x = "log₁₀ ratio",
      y = "Density"
    ) +
    theme_minimal(base_size = 14)

  ggsave(filename, plot = p, dpi = 300, width = 6, height = 4)
  return(p)
}


# Co-occurrence presence/absence test
run_presence_absence_test <- function(df_wide) {
  df_wide <- df_wide %>%
    mutate(
      smithii_present = ifelse(M_smithii > 1e-5, 1, 0),
      intestini_present = ifelse(M_intestini > 1e-5, 1, 0)
    )

  contingency_table <- table(df_wide$smithii_present, df_wide$intestini_present)

  # Ensure 2x2 matrix
  if (all(dim(contingency_table) == c(2, 2))) {
    fisher_test_result <- fisher.test(contingency_table)
  } else {
    warning("Contingency table is not 2x2. Fisher test skipped.")
    fisher_test_result <- NULL
  }

  return(list(
    table = contingency_table,
    fisher = fisher_test_result
  ))
}


```

# Extract table and stats

```{r}
# adults
adults_wide <- reshape_methanogen_abundances(adults_df)
plot_density_adults <- plot_log_ratio_density(adults_wide)
results_aults <- run_presence_absence_test(adults_wide)
results_aults$table
results_aults$fisher

# children
children_wide <- reshape_methanogen_abundances(children_df)
plot_density_children <- plot_log_ratio_density(children_wide)
results_chlidren <- run_presence_absence_test(children_wide)


results_aults$table
results_aults$fisher
results_chlidren$table
results_chlidren$fisher


plot_density_adults
plot_density_children
```



# assess correlations by country


```{r}
cor_methanogens_by_country_cooccur <- function(df, min_samples = 3) {
  df %>%
    select(Sample, OTU, Abundance, CurrCountry) %>%
    pivot_wider(
      names_from = OTU,
      values_from = Abundance,
      values_fill = 0
    ) %>%
    rename(
      M_smithii   = `s__Methanobrevibacter_A smithii`,
      M_intestini = `s__Methanobrevibacter_A smithii_A`
    ) %>%
    filter(M_smithii > 0 & M_intestini > 0) %>%
    group_by(CurrCountry) %>%
    summarise(
      n_samples = n(),
      spearman_rho = if (n() >= min_samples)
        cor(M_smithii, M_intestini, method = "spearman") else NA_real_,
      spearman_p = if (n() >= min_samples)
        cor.test(M_smithii, M_intestini, method = "spearman")$p.value else NA_real_,
      .groups = "drop"
    ) %>%
    mutate(
      spearman_p_adj = p.adjust(spearman_p, method = "BH")
    ) %>%
    filter(!is.na(spearman_rho))  # keep only valid correlations
}
res_cooccur_adults  <- cor_methanogens_by_country_cooccur(adults_df)
res_cooccur_children  <- cor_methanogens_by_country_cooccur(children_df)
res_cooccur_adults
res_cooccur_children
```










# Assessing mutual exclusion/co-occurence (species stats)

```{r}
summarize_methanogen_patterns_phyloseq <- function(physeq_obj, country_var = "CurrCountry") {
  require(dplyr)
  
  # Extract OTU table (samples as rows)
  otu <- as.data.frame(otu_table(physeq_obj))
  if (taxa_are_rows(physeq_obj)) otu <- t(otu)
  otu <- as.data.frame(otu)
  
  # rename species
  colnames(otu) <- dplyr::case_when(
    colnames(otu) == "s__Methanobrevibacter_A smithii"   ~ "M. smithii",
    colnames(otu) == "s__Methanobrevibacter_A smithii_A" ~ "M. intestini",
    TRUE ~ colnames(otu)
  )
  
  target <- c("M. smithii", "M. intestini")
  present <- intersect(target, colnames(otu))
  stopifnot(length(present) == 2)
  
  df <- otu[, present, drop = FALSE]
  df$Sample <- rownames(df)
  
  # Join metadata (country)
  meta <- as.data.frame(sample_data(physeq_obj))
  meta$Sample <- rownames(meta)
  df <- dplyr::left_join(df, meta[, c("Sample", country_var), drop = FALSE], by = "Sample")
  names(df)[names(df) == country_var] <- "CurrCountry"
  
  # Assign category
  df <- df %>%
    mutate(
      pattern = case_when(
        `M. smithii` > 0 & `M. intestini` == 0 ~ "M. smithii only",
        `M. smithii` == 0 & `M. intestini` > 0 ~ "M. intestini only",
        `M. smithii` > 0 & `M. intestini` > 0 & `M. smithii` > `M. intestini` ~ "Both: M. smithii dominant",
        `M. smithii` > 0 & `M. intestini` > 0 & `M. intestini` > `M. smithii` ~ "Both: M. intestini dominant",
        TRUE ~ "None"
      )
    )
  
  df
}

plot_methanogen_patterns <- function(df, title = "Methanogen occurrence and dominance", show_counts = TRUE) {
  require(dplyr)
  require(ggplot2)
  require(tidyr)
  
  # Define factor order for nice legend
  pattern_levels <- c("M. smithii only", "M. intestini only",
                      "Both: M. smithii dominant", "Both: M. intestini dominant", "None")
  df$pattern <- factor(df$pattern, levels = pattern_levels)
  
  sumdf <- df %>%
    count(CurrCountry, pattern, name = "n") %>%
    group_by(CurrCountry) %>%
    mutate(
      N = sum(n),
      prop = ifelse(N > 0, n / N, 0),
      Country_lab = paste0(CurrCountry, " (n=", N, ")"),
      label = ifelse(prop > 0,
                     if (show_counts) paste0(round(100 * prop, 1), "% (", n, ")")
                     else paste0(round(100 * prop, 1), "%"),
                     "")
    ) %>%
    ungroup()
  
  p <- ggplot(sumdf, aes(y = Country_lab, x = prop, fill = pattern)) +
    geom_col(position = "fill", width = 0.5) +
    geom_text(aes(label = label),
              position = position_fill(vjust = 0.5),
              color = "black",
              size = 7 / .pt) +
    scale_x_continuous(labels = scales::percent_format()) +
    scale_fill_manual(values = c(
      "M. smithii only"             = "#FFD700",   # gold
      "M. intestini only"           = "#6495ED",   # cornflower blue
      "Both: M. smithii dominant"   = "#66CD00", # chartreuse3
      "Both: M. intestini dominant" = "#4A6990FF", # dark blue
      "None"                        = "grey80"
    ), drop = FALSE) +
    labs(
      y = "Country (total n)",
      x = "Proportion of samples",
      fill = "Pattern",
      title = title,
      subtitle = "Includes single-species and co-occurrence dominance patterns"
    ) +
    theme_minimal(base_size = 7, base_family = "sans") +
    theme(legend.position = "top")
  
  return(p)
}
```

# generate plots

```{r}
# Adults
df_pat_adults <- summarize_methanogen_patterns_phyloseq(methanogens1)
p_adults <- plot_methanogen_patterns(df_pat_adults, title = "Adults: methanogen occurrence and dominance", show_counts = TRUE)

# Children
df_pat_children <- summarize_methanogen_patterns_phyloseq(methanogens2)
p_children <- plot_methanogen_patterns(df_pat_children, title = "Children: methanogen occurrence and dominance", show_counts = TRUE)

print(p_adults)
print(p_children)
```

# second function: excludes samples negative for both M. smithii & M. intestini

```{r}
# --- Summarize methanogen patterns per sample from phyloseq ---
summarize_methanogen_patterns_phyloseq <- function(physeq_obj, country_var = "CurrCountry") {
  require(dplyr)
  
  # Extract OTU table (samples as rows)
  otu <- as.data.frame(otu_table(physeq_obj))
  if (taxa_are_rows(physeq_obj)) otu <- t(otu)
  otu <- as.data.frame(otu)
  
  # Map GTDB names -> friendly names
  colnames(otu) <- dplyr::case_when(
    colnames(otu) == "s__Methanobrevibacter_A smithii"   ~ "M. smithii",
    colnames(otu) == "s__Methanobrevibacter_A smithii_A" ~ "M. intestini",
    TRUE ~ colnames(otu)
  )
  
  target <- c("M. smithii", "M. intestini")
  present <- intersect(target, colnames(otu))
  stopifnot(length(present) == 2)
  
  df <- otu[, present, drop = FALSE]
  df$Sample <- rownames(df)
  
  # Join metadata (country)
  meta <- as.data.frame(sample_data(physeq_obj))
  meta$Sample <- rownames(meta)
  df <- dplyr::left_join(df, meta[, c("Sample", country_var), drop = FALSE], by = "Sample")
  names(df)[names(df) == country_var] <- "CurrCountry"
  
  # Assign category (patterns)
  df <- df %>%
    mutate(
      pattern = case_when(
        `M. smithii` > 0 & `M. intestini` == 0 ~ "M. smithii only",
        `M. smithii` == 0 & `M. intestini` > 0 ~ "M. intestini only",
        `M. smithii` > 0 & `M. intestini` > 0 & `M. smithii` > `M. intestini` ~ "Both: M. smithii dominant",
        `M. smithii` > 0 & `M. intestini` > 0 & `M. intestini` > `M. smithii` ~ "Both: M. intestini dominant",
        TRUE ~ "None"
      )
    ) %>%
    filter(pattern != "None")   
  
  df
}


# --- Plot patterns with adaptive labels 
plot_methanogen_patterns <- function(df, title = "Methanogen occurrence and dominance", show_counts = TRUE) {
  require(dplyr)
  require(ggplot2)
  require(tidyr)
  
  # Define factor order for legend
  pattern_levels <- c(
    "M. smithii only",
    "M. intestini only",
    "Both: M. smithii dominant",
    "Both: M. intestini dominant"
  )
  df$pattern <- factor(df$pattern, levels = pattern_levels)
  
  # Define country order manually (Gabon on top)
  country_levels <- c("Vietnam", "Germany", "Gabon")
  
  # Summarize data
  sumdf <- df %>%
    count(CurrCountry, pattern, name = "n") %>%
    group_by(CurrCountry) %>%
    mutate(
      N = sum(n),
      prop = ifelse(N > 0, n / N, 0),
      Country_lab = paste0(CurrCountry, " (n=", N, ")"),
      label = ifelse(
        prop > 0,
        if (show_counts)
          paste0(round(100 * prop, 1), "% (", n, ")")
        else
          paste0(round(100 * prop, 1), "%"),
        ""
      )
    ) %>%
    ungroup()
  
  # Reorder factor levels for y-axis display
  sumdf <- sumdf %>%
    mutate(
      CurrCountry = factor(CurrCountry, levels = country_levels),
      Country_lab = factor(Country_lab, 
                           levels = paste0(country_levels, " (n=", 
                                           sapply(country_levels, function(x) {
                                             n_unique <- unique(sumdf$N[sumdf$CurrCountry == x])
                                             ifelse(length(n_unique) > 0, n_unique, NA)
                                           }), 
                                           ")"))
    )
  
  # Plot
  p <- ggplot(sumdf, aes(y = Country_lab, x = prop, fill = pattern)) +
    geom_col(position = "fill", width = 0.5) +
    geom_text(
      aes(label = label),
      position = position_fill(vjust = 0.5),
      color = "black",
      size = 7 / .pt
    ) +
    scale_x_continuous(labels = scales::percent_format()) +
    scale_fill_manual(
      values = c(
        "M. smithii only"             = "#FFD700",   # gold
        "M. intestini only"           = "#6495ED",   # cornflower blue
        "Both: M. smithii dominant"   = "#66CD00",   # chartreuse3
        "Both: M. intestini dominant" = "#4A6990FF"  # dark blue
      ),
      drop = FALSE
    ) +
    labs(
      y = "Country (total n)",
      x = "Proportion of samples",
      fill = "Pattern",
      title = title,
      subtitle = "Includes single-species and co-occurrence dominance patterns"
    ) +
    theme_minimal(base_size = 7, base_family = "sans") +
    theme(legend.position = "top")
  
  return(p)
}

```

# generate plot and save

```{r}
# Adults
df_pat_adults <- summarize_methanogen_patterns_phyloseq(methanogens1)
p_adults <- plot_methanogen_patterns(df_pat_adults, title = "Adults: methanogen occurrence and dominance", show_counts = TRUE)

# Children
df_pat_children <- summarize_methanogen_patterns_phyloseq(methanogens2)
p_children <- plot_methanogen_patterns(df_pat_children, title = "Children: methanogen occurrence and dominance", show_counts = TRUE)

print(p_adults)
print(p_children)

ggsave("Co-occurence_Ms_Mi_by_country_adults-Fig2.pdf", p_adults, width = 210, height = 120, units = "mm", dpi = 300)
ggsave("Co-occurence_Ms_Mi_by_country_adults-Fig2.svg", p_adults, width = 210, height = 120, units = "mm", dpi = 300, device = svglite::svglite)

ggsave("Co-occurence_Ms_Mi_by_country_children-Fig2.pdf", p_children, width = 210, height = 120, units = "mm", dpi = 300)
ggsave("Co-occurence_Ms_Mi_by_country_children-Fig2.svg", p_children, width = 210, height = 120, units = "mm", dpi = 300, device = svglite::svglite)

```

# write summary table to file

```{r}
summarize_methanogen_table <- function(physeq_obj, age_group, country_var = "CurrCountry") {
  require(dplyr)
  
  # Extract OTU table (samples as rows)
  otu <- as.data.frame(otu_table(physeq_obj))
  if (taxa_are_rows(physeq_obj)) otu <- t(otu)
  otu <- as.data.frame(otu)
  
  # Map GTDB names -> friendly names
  colnames(otu) <- dplyr::case_when(
    colnames(otu) == "s__Methanobrevibacter_A smithii"   ~ "M. smithii",
    colnames(otu) == "s__Methanobrevibacter_A smithii_A" ~ "M. intestini",
    TRUE ~ colnames(otu)
  )
  
  target <- c("M. smithii", "M. intestini")
  present <- intersect(target, colnames(otu))
  stopifnot(length(present) == 2)
  
  df <- otu[, present, drop = FALSE]
  df$Sample <- rownames(df)
  
  # Join metadata
  meta <- as.data.frame(sample_data(physeq_obj))
  meta$Sample <- rownames(meta)
  df <- dplyr::left_join(df, meta[, c("Sample", country_var), drop = FALSE], by = "Sample")
  names(df)[names(df) == country_var] <- "CurrCountry"
  
  # Assign category
  df <- df %>%
    mutate(
      pattern = case_when(
        `M. smithii` > 0 & `M. intestini` == 0 ~ "smithii_only",
        `M. smithii` == 0 & `M. intestini` > 0 ~ "intestini_only",
        `M. smithii` > 0 & `M. intestini` > 0 & `M. smithii` > `M. intestini` ~ "smithii_dom",
        `M. smithii` > 0 & `M. intestini` > 0 & `M. intestini` > `M. smithii` ~ "intestini_dom",
        TRUE ~ "neither"
      )
    )
  
  # Summarize per country
  out <- df %>%
    group_by(CurrCountry) %>%
    summarise(
      a_both = sum(pattern %in% c("smithii_dom", "intestini_dom")),
      b_intestini_only = sum(pattern == "intestini_only"),
      c_smithii_only = sum(pattern == "smithii_only"),
      d_neither = sum(pattern == "neither"),
      smithii_dom = sum(pattern == "smithii_dom"),
      intestini_dom = sum(pattern == "intestini_dom"),
      N = n(),
      .groups = "drop"
    ) %>%
    mutate(age_group = age_group) %>%
    select(CurrCountry, age_group, a_both, b_intestini_only, c_smithii_only,
           d_neither, N, smithii_dom, intestini_dom)
  
  return(out)
}
```

# generate table and write to file

```{r}
# Adults
tab_adults <- summarize_methanogen_table(methanogens1, age_group = "Adult")

# Infants
tab_children <- summarize_methanogen_table(methanogens2, age_group = "Infant")

# Combine
final_tab <- bind_rows(tab_adults, tab_children)


# Write to file
write.table(final_tab, "/data/methanogen_summary.txt", sep = "\t", quote = FALSE, row.names = FALSE)
```




## Microbiome Alpha diversity

# plot function

```{r}
plot_alpha_diversity_by_country <- function(physeq_obj, country_col = "CurrCountry") {
  # Calculate Observed and Shannon diversity
  alpha_df <- estimate_richness(physeq_obj, measures = c("Observed", "Shannon")) %>%
    rownames_to_column("SampleID") %>%
    left_join(as_tibble(sample_data(physeq_obj), rownames = "SampleID"), by = "SampleID")

  # Ensure consistent factor order for countries
  alpha_df[[country_col]] <- factor(alpha_df[[country_col]], levels = c("Gabon", "Germany", "Vietnam"))

  # Rename for consistency
  colnames(alpha_df)[colnames(alpha_df) == "Shannon"] <- "Shannon index"

  # Define comparisons
  comparisons <- list(
    c("Gabon", "Germany"),
    c("Gabon", "Vietnam"),
    c("Germany", "Vietnam")
  )

  # Colors
  country_colors <- c("Gabon" = "#ff0000", "Germany" = "#e7e70e", "Vietnam" = "#0b0bd3")

  # Function to generate plot for one metric
  make_alpha_plot <- function(metric_name, y_label) {
    df_metric <- alpha_df %>%
      select(SampleID, all_of(country_col), value = all_of(metric_name))

    y_max <- max(df_metric$value, na.rm = TRUE)
    y_positions <- seq(y_max * 1.05, length.out = length(comparisons), by = y_max * 0.05)

    ggplot(df_metric, aes_string(x = country_col, y = "value")) +
      geom_boxplot(outlier.shape = NA, fill = NA, color = "black") +
      #stat_summary(fun = median, geom = "text", aes(label = round(..y.., 3)), 
             #color = "blue", vjust = -0.5, size = 2.5) +
      geom_jitter(aes_string(color = country_col), width = 0.2, alpha = 0.5, size = 1) +
      stat_compare_means(
        comparisons = comparisons,
        method = "wilcox.test",
        label = "p.adj.format", # (or p.signif to label with stars/p.adj.format)
        p.adjust.method = "BH",
        tip.length = 0,
        vjust = 0.3,
        y.position = y_positions
      ) +
      scale_color_manual(values = country_colors) +
      labs(x = NULL, y = y_label) +
      theme_light(base_size = 7, base_family = "sans") +
      theme(
        legend.position = "none",
        axis.text = element_text(size = 7, color = "black"),
        axis.text.x = element_text(size = 7, color = "black"), # (or axis.text.x = element_text(angle = 45, hjust = 1))
        axis.title = element_text(size = 7, color = "black"),
        plot.title = element_blank()
      )
  }

  # Return both plots in a list
  list(
    Observed = make_alpha_plot("Observed", "Observed richness"),
    Shannon = make_alpha_plot("Shannon index", "Shannon diversity index")
  )
}

```



# Alpha diversity all microbes (plot observed richness)

```{r}
library(ggplot2)
alpha_plots_adults_all <- plot_alpha_diversity_by_country(ps_adults_raw)
alpha_plots_children_all <- plot_alpha_diversity_by_country(ps_children_raw)

alpha_plots_adults_all$Observed

alpha_plots_children_all$Observed


ggsave("Observed-richness-children-extended-data-Fig4.pdf", alpha_plots_children_all$Observed, width = 95, height = 80, units = "mm", dpi = 300)
ggsave("Observed-richness-children-extended-data-Fig4.svg", alpha_plots_children_all$Observed, width = 95, height = 80, units = "mm", device = svglite::svglite, dpi = 300)
ggsave("Observed-richness-adults-extended-data-Fig4.pdf", alpha_plots_adults_all$Observed, width = 95, height = 80, units = "mm", dpi = 300)
ggsave("Observed-richness-adults-extended-data-Fig4.svg", alpha_plots_adults_all$Observed, width = 95, height = 80, units = "mm", device = svglite::svglite, dpi = 300)

```

# Alpha diversity (Qiime2 rarefied to 500k reads: plot Shannon entropy and Faith's PD)

```{r}
metadat_adults = read.table("/data/filtered_adult_metadata.txt", header = TRUE, sep = "\t")

metadat_children = read.table("/data/filtered_child_metadata.txt", header = TRUE, sep = "\t")

```

# plot from obtained shannon_entropy & faith_pd in metadata tables

```{r}
plot_alpha_diversity_from_metadata <- function(meta_df, country_col = "CurrCountry") {
  require(ggplot2)
  require(dplyr)
  require(ggpubr)
  
  # Ensure country order is consistent
  meta_df[[country_col]] <- factor(meta_df[[country_col]], levels = c("Gabon", "Germany", "Vietnam"))
  
  # Define pairwise comparisons
  comparisons <- list(
    c("Gabon", "Germany"),
    c("Gabon", "Vietnam"),
    c("Germany", "Vietnam")
  )
  
  # Define colors
  country_colors <- c("Gabon" = "#ff0000", "Germany" = "#e7e70e", "Vietnam" = "#0b0bd3")
  
  # --- internal plotting function ---
  make_alpha_plot <- function(metric_col, y_label) {
    df_metric <- meta_df %>%
      select(all_of(country_col), value = all_of(metric_col)) %>%
      filter(!is.na(value))
    
    y_max <- max(df_metric$value, na.rm = TRUE)
    y_positions <- seq(y_max * 1.05, by = y_max * 0.05, length.out = length(comparisons))
    
    ggplot(df_metric, aes_string(x = country_col, y = "value", color = country_col)) +
      geom_boxplot(outlier.shape = NA, fill = NA, color = "black", linewidth = 0.2) +
      geom_jitter(width = 0.2, alpha = 0.6, size = 0.8) +
      #stat_summary(fun = median, geom = "text", 
                   #aes(label = round(..y.., 2)), 
                   #color = "black", vjust = -0.5, size = 7 / .pt) +
      stat_compare_means(
        comparisons = comparisons,
        method = "wilcox.test",
        label = "p.adj.format", 
        p.adjust.method = "BH",
        tip.length = 0,
        vjust = 0.3,
        size = 7 / .pt,
        y.position = y_positions
      ) +
      scale_color_manual(values = country_colors) +
      labs(
        x = NULL,
        y = y_label,
        color = "Country"
      ) +
      theme_light(base_size = 7, base_family = "sans") +
      theme(
        legend.position = "none",
        axis.text = element_text(size = 7, color = "black"),
        axis.title = element_text(size = 7, color = "black"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.2)
      )
  }
  
  # Return a list of plots
  list(
    Shannon = make_alpha_plot("shannon_entropy", "Shannon diversity index"),
    FaithPD = make_alpha_plot("faith_pd", "Faith’s phylogenetic diversity")
  )
}
```

# generate and save plots

```{r}
# Adults
alpha_plots_adults <- plot_alpha_diversity_from_metadata(metadat_adults)
alpha_plots_adults$Shannon
alpha_plots_adults$FaithPD

# Children
alpha_plots_children <- plot_alpha_diversity_from_metadata(metadat_children)
alpha_plots_children$Shannon
alpha_plots_children$FaithPD

ggsave("Faiths-PD-children-extended-data-Fig4.pdf", alpha_plots_children$FaithPD, width = 95, height = 80, units = "mm", dpi = 300)
ggsave("Faiths-PD-children-extended-data-Fig4.svg", alpha_plots_children$FaithPD, width = 95, height = 80, units = "mm", device = svglite::svglite, dpi = 300)
ggsave("Faiths-PD-adults-extended-data-Fig4.pdf", alpha_plots_adults$FaithPD, width = 95, height = 80, units = "mm", dpi = 300)
ggsave("Faiths-PD-adults-extended-data-Fig4.svg", alpha_plots_adults$FaithPD, width = 95, height = 80, units = "mm", device = svglite::svglite, dpi = 300)

```

# correlation of alpha diversity (Shannon entropy) and M. smithii/M. intestini abundance

```{r}
adults_df
children_df
metadat_adults
metadat_children
```

```{r}

### 1. Define methanogen names 


ms_name <- "s__Methanobrevibacter_A smithii"
mi_name <- "s__Methanobrevibacter_A smithii_A"


### 2. Function to extract MS + MI from abundance table


extract_methanogens <- function(abund_df) {
  
  abund_df %>%
    select(ParticipantID, OTU, Abundance) %>%
    
    # Keep only the two methanogens
    filter(OTU %in% c(ms_name, mi_name)) %>%
    
    # Pivot wider so each becomes a column
    pivot_wider(
      names_from = OTU,
      values_from = Abundance,
      values_fill = 0
    ) %>%
    
    # Rename columns
    rename(
      MS_abundance = !!ms_name,
      MI_abundance = !!mi_name
    )
}


### 3. Apply to adults and children


adult_meth <- extract_methanogens(adults_df)
child_meth <- extract_methanogens(children_df)


### 4. Join with metadata tables


adults_meta_final <- metadat_adults %>%
  left_join(adult_meth, by = "ParticipantID") %>%
  mutate(Age_group = "Adult")

children_meta_final <- metadat_children %>%
  left_join(child_meth, by = "ParticipantID") %>%
  mutate(Age_group = "Infant")


### 5. Combine adults + infants datasets


combined_meta <- bind_rows(adults_meta_final, children_meta_final)


### 6. Order columns 


combined_meta <- combined_meta %>%
  relocate(Age_group, ParticipantID, MS_abundance, MI_abundance)

```

# fixed-effect linear models

```{r}
# 1. Combined model (adults + infants)

model_all <- lm(
  shannon_entropy ~ MS_abundance + Age_group + CurrCountry,
  data = combined_meta 
)
model_all

# 2. Adults only

model_adults <- lm(
  shannon_entropy ~ MS_abundance + Age_years + samplingLocation,
  data = combined_meta  %>% filter(Age_group == "Adult")
)
model_adults

# 3. Infants only
model_infants <- lm(
  shannon_entropy ~ MS_abundance + Age_months + samplingLocation,
  data = combined_meta %>% filter(Age_group == "Infant")
)
model_infants
```

# generate tidied output tables

```{r}
library(dplyr)
library(broom)
library(readr)
library(writexl)

run_methanogen_models <- function(data, abundance_col) {
  
  formula_all <- as.formula(
    paste("shannon_entropy ~", abundance_col, "+ Age_group + CurrCountry")
  )
  
  formula_adults <- as.formula(
    paste("shannon_entropy ~", abundance_col, "+ Age_years + samplingLocation")
  )
  
  formula_infants <- as.formula(
    paste("shannon_entropy ~", abundance_col, "+ Age_months + samplingLocation")
  )
  
  # Fit models
  model_all <- lm(formula_all, data = data)
  model_adults <- lm(formula_adults, data = data %>% filter(Age_group == "Adult"))
  model_infants <- lm(formula_infants, data = data %>% filter(Age_group == "Infant"))

  # Return tidy results with model labels
  bind_rows(
    tidy(model_all) %>% mutate(Model = "Combined"),
    tidy(model_adults) %>% mutate(Model = "Adults"),
    tidy(model_infants) %>% mutate(Model = "Infants")
  ) %>%
    relocate(Model)
}

# run functions
results_MS <- run_methanogen_models(combined_meta, "MS_abundance")
results_MI <- run_methanogen_models(combined_meta, "MI_abundance")

# Combined table
all_results <- bind_rows(
  results_MS %>% mutate(Methanogen = "M. smithii"),
  results_MI %>% mutate(Methanogen = "M. intestini")
) %>% 
  relocate(Methanogen, Model)

all_results

# write to table
write_csv(all_results, "methanogen_alpha_diversity_regression_results.csv")

```


# extract and tidy results stats

```{r}
library(broom)
library(dplyr)

fit_and_compare <- function(meta, meth_var) {
  
  # Combined model
  mod_all <- lm(
    as.formula(paste("shannon_entropy ~", meth_var, "+ Age_group + CurrCountry")),
    data = meta
  )
  
  # Adults only
  adult_data <- meta %>%
    filter(Age_group == "Adult") %>%
    select(shannon_entropy, all_of(meth_var), Age_years, samplingLocation) %>%
    na.omit()
  
  mod_ad_full <- lm(
    as.formula(paste("shannon_entropy ~", meth_var, "+ Age_years + samplingLocation")),
    data = adult_data
  )
  mod_ad_red <- lm(
    as.formula(paste("shannon_entropy ~", meth_var, "+ samplingLocation")),
    data = adult_data
  )
  
  anova_ad <- anova(mod_ad_red, mod_ad_full)
  
  # Infants only
  infant_data <- meta %>%
    filter(Age_group == "Infant") %>%
    select(shannon_entropy, all_of(meth_var), Age_months, samplingLocation) %>%
    na.omit()
  
  mod_in_full <- lm(
    as.formula(paste("shannon_entropy ~", meth_var, "+ Age_months + samplingLocation")),
    data = infant_data
  )
  
  # Combine tidy results
  cis <- bind_rows(
    broom::tidy(mod_all, conf.int = TRUE),
    broom::tidy(mod_ad_full, conf.int = TRUE),
    broom::tidy(mod_ad_red, conf.int = TRUE),
    broom::tidy(mod_in_full, conf.int = TRUE)
  ) %>%
    mutate(Model = c("Combined","Adult_full","Adult_reduced","Infant_full")[rep(1:4, times = c(
      nrow(broom::tidy(mod_all)), 
      nrow(broom::tidy(mod_ad_full)), 
      nrow(broom::tidy(mod_ad_red)), 
      nrow(broom::tidy(mod_in_full))
    ))])
  
  return(list(tidy = cis, anova_adult = anova_ad))
}

res_MS <- fit_and_compare(combined_meta, "MS_abundance")
res_MI <- fit_and_compare(combined_meta, "MI_abundance")

res_MS
res_MI

write_csv(res_MS$tidy, "MS_alpha_diversity_regression_results.csv")
write_csv(res_MS$tidy, "MI_alpha_diversity_regression_results.csv")
write.table(res_MS$tidy, file = "/data/MS_alpha_diversity_regression_results.csv", row.names = FALSE, sep = "\t")
write.table(res_MI$tidy, file = "/data/MI_alpha_diversity_regression_results.csv", row.names = FALSE, sep = "\t")

```

