---
title: "01_Methanogen_prevalence_abundance"
author: "Mirabeau Mbong"
date: "2025-06-26"
output:
  html_notebook:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
---
This script is used to generate plots in Fig.1 and Fig.S2 of the associated paper

```{r, setup, include=FALSE}
knitr::opts_chunk$set(dev = "pdf",
                      dpi = 300,
                      echo = FALSE,
                      cache = TRUE)

```

# load libraries

```{r}
# Load required libraries
library(phyloseq)
library(microbiome)
library(tidyverse)
library(ggpubr)
library(patchwork)
library(ggpubr)
library(vegan)
library(tibble)
library(showtext)
library(showtextdb)
library(ggtext)
library(ggplot2)
library(tidyr)
library(dplyr)
```

```{r}
# Check working directory (should be repo root)
message("Current working directory: ", getwd())
```

# Functions: data processing I (compute prevalence across all samples)

```{r}
# ---------- Function to load Bracken + metadata and create phyloseq ----------
create_phyloseq <- function(bracken_file, metadata_file) {
  bracken <- read.table(bracken_file, header = TRUE, sep = "\t")
  meta <- read.table(metadata_file, header = TRUE, sep = "\t")

  # Process OTU table
  OTU <- bracken %>%
    dplyr::select(-c(taxonomy_id, taxonomy_lvl, taxonomy, taxIDs)) %>%
    dplyr::select(-contains('frac')) %>%
    dplyr::rename_with(~ sub("[_]num$", "", .x), ends_with("_num"))

  rownames(OTU) <- OTU$name
  OTU <- OTU %>% dplyr::select(-name)

  # Process taxonomy
  tax <- bracken %>%
    dplyr::select(name, taxonomy) %>%
    tidyr::separate(taxonomy,
                    into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
                    sep = ";", fill = "right", extra = "drop")
  rownames(tax) <- tax$name
  tax <- tax %>% dplyr::select(-name)

  # Process metadata
  meta$sample <- meta$ParticipantID
  rownames(meta) <- meta$sample
  meta <- meta %>% dplyr::select(-sample)

  # Create phyloseq object
  ps <- phyloseq(
    otu_table(as.matrix(OTU), taxa_are_rows = TRUE),
    tax_table(as.matrix(tax)),
    sample_data(meta)
  )
  return(ps)
}

# ---------- Usage of phyloseq function ----------

# ---------- Load data ----------
ps_adults_raw <- create_phyloseq("data/bracken_output_adults.tsv", "data/filtered_adult_metadata.txt")
ps_children_raw <- create_phyloseq("data/bracken_output_children.tsv", "data/filtered_child_metadata.txt")


# ---------- Function to filter and clean methanogens ----------

# Filter by phylum: Methanobacteriota
filter_methanogen_phylum <- function(ps) {
  subset_taxa(ps, Phylum == "p__Methanobacteriota")
}

# ---------- Clean and filter Methanobrevibacter_A genus ----------
filter_clean_methanobrevibacter_A <- function(ps) {
  ps_meth <- subset_taxa(ps, Genus == "g__Methanobrevibacter_A")
  
  tax_df <- as.data.frame(tax_table(ps_meth))
  tax_df$Species[is.na(tax_df$Species)] <- ""
  keep_taxa <- rownames(tax_df)[!grepl("sp", tax_df$Species)]
  ps_clean <- prune_taxa(keep_taxa, ps_meth)
  
  tax <- tax_table(ps_clean)
  tax[, "Genus"] <- gsub("^g__", "", tax[, "Genus"])
  tax[, "Species"] <- gsub("^s__", "", tax[, "Species"])
  tax[, "Species"] <- gsub("Methanobrevibacter_A smithii_A", 
                           "Methanobrevibacter_A intestini", tax[, "Species"])
  tax_table(ps_clean) <- tax
  
  return(ps_clean)
}

# ---------- usage of fitering function ----------

# Adults
methanogens1 <- filter_methanogen_phylum(ps_adults_raw)
methanobrevibacter1 <- filter_clean_methanobrevibacter_A(methanogens1)

# Children
methanogens2 <- filter_methanogen_phylum(ps_children_raw)
methanobrevibacter2 <- filter_clean_methanobrevibacter_A(methanogens2)

# ---------- All samples prevalence Computation ----------
compute_prevalence <- function(ps_obj, total_samples) {
  # Calculate prevalence per taxon
  prevalence_counts <- apply(
    X = otu_table(ps_obj),
    MARGIN = ifelse(taxa_are_rows(ps_obj), 1, 2),
    FUN = function(x) sum(x > 0)
  )
  
  # Build data.frame with prevalence, abundance, and taxonomy
  prev_df <- data.frame(
    Prevalence = prevalence_counts,
    TotalAbundance = taxa_sums(ps_obj),
    tax_table(ps_obj)
  )
  
  # Calculate percent prevalence
  prev_df <- prev_df %>%
    dplyr::mutate(prevalence = (Prevalence / total_samples) * 100)
  
  return(prev_df)
}

# ---------- Usage of prevalence function ----------

# For adults (510 samples)
prev_adults <- compute_prevalence(methanogens1, total_samples = 510)

# For children (394 samples)
prev_children <- compute_prevalence(methanogens2, total_samples = 394)

```

# write tables


```{r}
write.table(prev_adults,file = "Prevalence_methanogens-adults.txt", row.names = FALSE, sep = "\t")
write.table(prev_children,file = "Prevalence_methanogens-children.txt", row.names = FALSE, sep = "\t")

```

# Functions: data processing II (compute country prevalence)


```{r}
# ---------- compute prevalence function ----------

compute_prevalence_by_group <- function(physeq_obj, group_var = "CurrCountry", abundance_threshold = 1e-1) {
  # Filter abundant taxa
  physeq_filtered <- filter_taxa(physeq_obj, function(x) mean(x) > abundance_threshold, TRUE)
  
  # Melt OTU table
  otu <- otu_table(physeq_filtered)
  if (!taxa_are_rows(physeq_filtered)) otu <- t(otu)
  otu_df <- as.data.frame(otu) %>% rownames_to_column("Taxon") %>%
    pivot_longer(-Taxon, names_to = "SampleID", values_to = "Abundance")
  
  # Add metadata
  sample_meta <- data.frame(sample_data(physeq_filtered)) %>% rownames_to_column("SampleID")
  otu_long <- left_join(otu_df, sample_meta, by = "SampleID")
  
  # Add taxonomy
  tax_df <- as.data.frame(tax_table(physeq_filtered)) %>% rownames_to_column("Taxon")
  otu_long <- left_join(otu_long, tax_df, by = "Taxon")
  
  # Compute prevalence
  prevalence_df <- otu_long %>%
    group_by(!!sym(group_var), Taxon, Genus, Species) %>%
    summarise(prevalence = mean(Abundance > 0) * 100, .groups = "drop") %>%
    mutate(
      Genus = str_remove(Genus, "^g__"),
      Species = str_remove(Species, "^s__"),
      Species = str_replace_all(Species, "Methanobrevibacter_A", "M.")
    )
  
  names(prevalence_df)[1] <- group_var  # Ensure group var is named properly
  return(prevalence_df)
}

# ---------- compute abundance function ----------

prepare_methanogen_abundance_df <- function(physeq_obj) {
  relabund_obj <- transform_sample_counts(physeq_obj, function(x) x / sum(x))
  df_long <- psmelt(relabund_obj)

  species_df <- df_long %>%
    filter(OTU %in% c("s__Methanobrevibacter_A smithii", "s__Methanobrevibacter_A smithii_A")) %>%
    mutate(
      Species = case_when(
        OTU == "s__Methanobrevibacter_A smithii" ~ "M. smithii",
        OTU == "s__Methanobrevibacter_A smithii_A" ~ "M. intestini"
      ),
      Abundance = Abundance + 1e-7  # Pseudo-count for log scale
    )

  return(species_df)
}

```


#  Plotting functions


```{r}
# ---------- Custom color schemes ----------
country_colors <- c(Gabon = "#ff0000", Germany = "#e7e70e", Vietnam = "#0b0bd3")
species_colors <- c("Methanobrevibacter_A smithii" = "#1f77b4", "Methanobrevibacter_A intestini" = "#ff7f0e")
location_colors <- c("Lambaréné" = "skyblue", "Ikobey" = "orange")

plot_genus_prevalence <- function(prev_df) {
  prev_df_clean <- prev_df %>%
    filter(prevalence > 0) %>%  # Remove zero prevalence to avoid log10(0)
    mutate(
      Genus = gsub("^g__", "", Genus),
      Genus = gsub("^s__", "", Genus),
      Genus = ifelse(Genus == "Methanobrevibacter_A smithii_A", "Methanobrevibacter_A intestini", Genus)
    )

  # Compute maximum prevalence and relevel Genus by descending max prevalence
  max_prev_order <- prev_df_clean %>%
    group_by(Genus) %>%
    summarise(max_prev = max(prevalence, na.rm = TRUE)) %>%
    arrange(desc(max_prev)) %>%
    filter(!is.na(Genus))

  # Reorder factor levels BEFORE plotting
  prev_ordered <- prev_df_clean %>%
    filter(Genus %in% max_prev_order$Genus) %>%
    mutate(Genus = factor(Genus, levels = max_prev_order$Genus))

  # Plot
  ggplot(prev_ordered, aes(x = Genus, y = prevalence)) +
    geom_segment(aes(xend = Genus, y = 0.1, yend = prevalence), color = "grey60", linewidth = 0.7) +
    geom_point(color = "black", size = 2) +
    scale_y_log10(
      limits = c(0.1, 100),
      breaks = c(0.1, 1, 5, 10, 25, 50, 100),
      labels = c("0.1", "1", "5", "10", "25", "50", "100")
    ) +
    scale_x_discrete(labels = function(x) parse(text = paste0("italic('", x, "')"))) +
    labs(x = "", y = "Prevalence (log scale %)") +
    theme_light(base_size = 7, base_family = "Helvetica") +
    theme(
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
      strip.text = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 7),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()
    )
}

plot_species_prevalence <- function(prev_df, genus_filter = "Methanobrevibacter_A", min_prevalence = 0.1) {
  prev_species <- prev_df %>%
    mutate(
      Genus = gsub("^g__", "", Genus),
      Species = gsub("^s__", "", Species),
      Species = gsub("^Methanobrevibacter_A smithii_A$", "Methanobrevibacter_A intestini", Species)
    ) %>%
    filter(
      Genus == genus_filter,
      !grepl("sp", Species),           # Remove MAGs
      prevalence > min_prevalence
    ) %>%
    mutate(
      Species = gsub(paste0("^", genus_filter), "M.", Species),
      Species = factor(Species, levels = rev(Species[order(prevalence)]))
    )
  
  ggplot(prev_species, aes(x = Species, y = prevalence)) +
    geom_segment(aes(xend = Species, y = 0, yend = prevalence), color = "grey60", linewidth = 0.7) +
    geom_point(color = "black", size = 2) +
    scale_y_continuous(
      limits = c(0, max(prev_species$prevalence, na.rm = TRUE) + 5),
      breaks = seq(0, 100, 10),
      labels = function(x) x
    ) +
    scale_x_discrete(labels = function(x) parse(text = paste0("italic('", x, "')"))) +
    labs(x = "", y = "Prevalence (%)") +
    theme_light(base_size = 7, base_family = "Helvetica") +
    theme(
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
      strip.text = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 7),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()
    )
}

# ---------- plot genus prevalence function ----------

plot_genus_max_prevalence_bar <- function(prev_df, group_var = "CurrCountry") {
  genus_df <- prev_df %>%
    filter(!is.na(Genus), prevalence > 0) %>%
    group_by(Genus, !!sym(group_var)) %>%
    summarise(prevalence = max(prevalence), .groups = "drop") %>%
    mutate(Genus = factor(Genus, levels = rev(unique(Genus[order(prevalence, decreasing = TRUE)]))))
  
  p <- ggplot(genus_df, aes(x = Genus, y = prevalence, fill = !!sym(group_var))) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.6) +
    coord_flip() +
    scale_y_sqrt(
      breaks = c(0, 1, 5, 10, 25, 50, 75, 100),
      labels = function(x) x

    ) +
    scale_fill_manual(name = "", values = country_colors) +
    scale_x_discrete(labels = function(x) parse(text = paste0("italic('", x, "')"))) +
    labs(x = "", y = "Max genus prevalence", fill = group_var) +
    theme_light(base_size = 7, base_family = "Helvetica") +
    theme(
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
      strip.text = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 7),
      legend.position = "none")

  return(p)
}

# ---------- plot species prevalence function ----------

plot_methanobrevibacter_species_bar <- function(prev_df, group_var = "CurrCountry", min_prevalence = 0.1) {
  meth_df <- prev_df %>%
    filter(Genus == "Methanobrevibacter_A", !str_detect(Species, "sp[0-9]+"), prevalence > min_prevalence) %>%
    mutate(
      Species = str_replace(Species, "M\\. smithii_A", "M. intestini")
    )
  
  if (nrow(meth_df) == 0) {
    warning("No eligible Methanobrevibacter species found.")
    return(NULL)
  }

  species_levels <- meth_df %>%
    group_by(Species) %>%
    summarise(median_prev = median(prevalence, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(median_prev)) %>%
    pull(Species)
  
  meth_df$Species <- factor(meth_df$Species, levels = rev(species_levels))

  ymax <- ceiling(max(meth_df$prevalence, na.rm = TRUE) / 10) * 10

  p <- ggplot(meth_df, aes(x = Species, y = prevalence, fill = !!sym(group_var))) +
    geom_bar(stat = "identity", position = position_dodge2(width = 0.7, preserve = "single"), width = 0.6) +
    coord_flip() +
    scale_y_continuous(
      limits = c(0, ymax),
      breaks = seq(0, ymax, 10),
      labels = function(x) x  
    ) +
    scale_fill_manual(name = "", values = country_colors) +
    scale_x_discrete(labels = function(x) parse(text = paste0("italic('", x, "')"))) +
    labs(x = "", y = "Prevalence (%)", fill = group_var) +  
    theme_light(base_size = 7, base_family = "Helvetica") +
    theme(
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
      strip.text = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 7),
      legend.position = "none"
    )

  return(p)
}

# ---------- plot abundance by country ----------

plot_methanogen_abundance <- function(species_df, group_var, comparisons_list,
                                            xlab = "", country_colors = NULL) {
  # Filter out NA or zero values for plotting on log10 scale
  filtered_df <- species_df %>%
    filter(!is.na(Abundance), Abundance > 0)

  # Determine x-axis label
  x_label <- if (is.null(xlab) || xlab == "") group_var else xlab

  # Custom labeller for italic species names
  species_labeller <- function(name) {
    as.expression(bquote(italic(.(name))))
  }

  # Split by species and generate one plot per species
  plots <- filtered_df %>%
    split(.$Species) %>%
    lapply(function(df_species) {
      species_name <- unique(df_species$Species)

      ggplot(df_species, aes_string(x = group_var, y = "Abundance * 100", color = group_var)) +
        geom_boxplot(outlier.shape = NA, fill = NA, color = "black", alpha = 0.8) +
        geom_jitter(width = 0.2, alpha = 0.6, size = 1) +
        stat_compare_means(label = "p.signif", method = "wilcox.test", comparisons = comparisons_list, tip.length = 0, vjust = 0.3) +
        scale_y_log10(labels = scales::label_number(accuracy = 0.0001)) +
        scale_color_manual(values = country_colors, na.translate = FALSE) +
        labs(
          y = "Relative abundance (log10 %)",
          x = NULL
          #title = species_labeller(species_name)
        ) +
        theme_light(base_size = 7, base_family = "Helvetical") +
        theme(
          plot.title = element_text(size = 8, hjust = 0.5),
          axis.text = element_text(size = 7),
          axis.title = element_text(size = 7),
          legend.position = "none"
        )
    })

  return(plots)
}


# ---------- plot abundance by country ----------

```

# Manuscript Figure (data prevalence: Supplementary Fig.02)

```{r}
# Adults
plg_adults = plot_genus_prevalence(prev_adults)
pls_adults = plot_species_prevalence(prev_adults)

# Children
plg_children = plot_genus_prevalence(prev_children)
pls_children = plot_species_prevalence(prev_children)

ggsave("genus_prevalence_adults-all-samples.pdf", plg_adults, width = 380, height = 180, units = "mm", dpi = 300)
ggsave("genus_prevalence_adults-all-samples.svg", plg_adults, width = 380, height = 180, units = "mm", dpi = 300, device = svglite::svglite)

plg_adults
pls_adults
plg_children
pls_children
```

```{r}

```


# Manuscript Figure (prevalence: Fig.1 & Supplementary Fig.02)


```{r}
# By country (adults):
prev_by_country <- compute_prevalence_by_group(methanogens1, group_var = "CurrCountry")
p_genus_country <- plot_genus_max_prevalence_bar(prev_by_country, group_var = "CurrCountry")
p_species_country <- plot_methanobrevibacter_species_bar(prev_by_country, group_var = "CurrCountry")

# By country (children):
prev_by_country1 <- compute_prevalence_by_group(methanogens2, group_var = "CurrCountry")
p_genus_country1 <- plot_genus_max_prevalence_bar(prev_by_country1, group_var = "CurrCountry")
p_species_country1 <- plot_methanobrevibacter_species_bar(prev_by_country1, group_var = "CurrCountry")

ggsave("species_prevalence_children-by-country.pdf", p_species_country1, width = 65, height = 45, units = "mm", dpi = 300)
ggsave("species_prevalence_children-by-country.svg", p_species_country1, width = 110, height = 75, units = "mm", dpi = 300, device = svglite::svglite)


p_genus_country
p_species_country
p_genus_country1
p_species_country1
```

# Manuscript Figure (abundance by country: Fig.1) usage  & sampling locations

```{r}
# Prepare data
adults_df <- prepare_methanogen_abundance_df(ps_adults_raw)
children_df <- prepare_methanogen_abundance_df(ps_children_raw)

# Color palette (for countries)
country_colors <- c("Gabon" = "#ff0000", "Germany" = "#e7e70e", "Vietnam" = "#0b0bd3")

# Comparisons
comparisons_list <- list(c("Gabon", "Germany"), c("Gabon", "Vietnam"), c("Germany", "Vietnam"))

# Call the function
species_plots_adults <- plot_methanogen_abundance(adults_df, group_var = "CurrCountry",
                                         comparisons_list = comparisons_list,
                                         xlab = "", country_colors = country_colors)

species_plots_children <- plot_methanogen_abundance(children_df, group_var = "CurrCountry",
                                         comparisons_list = comparisons_list,
                                         xlab = "", country_colors = country_colors)

# Show individual plots
species_plots_adults[["M. smithii"]]
species_plots_adults[["M. intestini"]]
species_plots_children[["M. smithii"]]
species_plots_children[["M. intestini"]]

ggsave("M_intestini_abundance_adults.pdf", species_plots_adults[["M. intestini"]], width = 50, height = 60, units = "mm", dpi = 300)
ggsave("M_intestini_abundance_adults.svg", species_plots_adults[["M. intestini"]], width = 80, height = 70, units = "mm", dpi = 300, device = svglite::svglite)

```

```{r}

```


#  Manuscript Figure (sampling locations: Fig.1)

```{r}
# Gabon only: filter and plot by sampling location
adults_gabon <- adults_df %>% filter(CurrCountry == "Gabon")
children_gabon <- children_df %>% filter(CurrCountry == "Gabon")
gabon_comparisons <- list(c("Lambarene", "Ikobey"))
location_colors <- c("Lambarene" = "skyblue", "Ikobey" = "orange")

#adults
species_plots_adults_loc <- plot_methanogen_abundance(
  adults_gabon,
  group_var = "samplingLocation",
  comparisons_list = gabon_comparisons,
  xlab = "",
  country_colors = location_colors
)

# children
species_plots_children_loc <- plot_methanogen_abundance(
  children_gabon,
  group_var = "samplingLocation",
  comparisons_list = gabon_comparisons,
  xlab = "",
  country_colors = location_colors
)

# Show individual plots
species_plots_adults_loc[["M. smithii"]]
species_plots_adults_loc[["M. intestini"]]
species_plots_children_loc[["M. smithii"]]
species_plots_children_loc[["M. intestini"]]

ggsave("M_intestini_abundance_adults-Gabon.pdf", species_plots_adults_loc[["M. intestini"]], width = 50, height = 60, units = "mm", dpi = 300)
ggsave("M_smithii_abundance_adults-Gabon.svg", species_plots_adults_loc[["M. smithii"]], width = 80, height = 70, units = "mm", dpi = 300, device = svglite::svglite)

```

```{r}

```


# Adults Vs Children abundance comparisons

```{r}

# Prepare adult and child subsets
min_abundance_adults <- 1e-06
adults_df_subset <- adults_df %>%
  filter(Abundance > min_abundance_adults) %>%
  select(Sample, Abundance, samplingLocation, CurrCountry, Species) %>%
  mutate(Dataset = "Adults")

children_df_subset <- children_df %>%
  select(Sample, Abundance, samplingLocation, CurrCountry, Species) %>%
  mutate(Dataset = "Children")

# Combine
combined_df <- bind_rows(adults_df_subset, children_df_subset) %>%
  mutate(
    CountryGroup = paste(CurrCountry, Dataset),
    LocationGroup = paste(samplingLocation, Dataset)
  )
```


# plot function


```{r}
plot_methanogen_abundance_grouped <- function(df, group_var, comparisons_list, color_by = NULL) {
  # Add ColorGroup column based on group_var
  df <- df %>%
    mutate(ColorGroup = dplyr::case_when(
      group_var == "CountryGroup" ~ CurrCountry,
      group_var == "LocationGroup" ~ samplingLocation,
      TRUE ~ NA_character_
    ))

  # Define color map
  if (color_by == "country") {
    color_map <- country_colors
  } else if (color_by == "location") {
    color_map <- location_colors
  } else {
    color_map <- NULL
  }

  # Custom labeller for italic species names
  species_labeller <- function(labels) {
    sapply(labels, function(x) paste0("italic('", x, "')"))
  }

  species_list <- unique(df$Species)
  plot_list <- list()

  for (species_name in species_list) {
    df_species <- df %>% filter(Species == species_name)

    y.max <- max(df_species$Abundance * 100, na.rm = TRUE)
    y.position <- seq(y.max * 1.1, length.out = length(comparisons_list), by = y.max * 0.1)

    p <- ggplot(df_species, aes_string(x = group_var, y = "Abundance * 100")) +
      geom_boxplot(outlier.shape = NA, fill = NA, color = "black", alpha = 0.8) +
      geom_jitter(aes(color = ColorGroup), width = 0.3, alpha = 0.3) +
      scale_y_log10(labels = scales::label_number(accuracy = 0.0001)) +
      stat_compare_means(
        label = "p.signif",
        method = "wilcox.test",
        comparisons = comparisons_list,
        tip.length = 0,
        vjust = 0.3
      ) +
      scale_color_manual(values = color_map) +
      labs(
        y = "Relative abundance (log10 %)",
        x = NULL  # removes redundant x-axis label
      ) +
      theme_light(base_size = 7, base_family = "Helvetica") +
      theme(
        strip.text = element_text(size = 7),
        axis.text = element_text(size = 7),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.title = element_text(size = 7),
        text = element_text(size = 7, colour = "black"),
        legend.position = "none"
      )

    plot_list[[species_name]] <- p
  }

  return(plot_list)
}

```

# Manuscript Figure (Adults Vs Chilcdren abundance comparisons: supplementary Fig.02)


```{r}
country_comparisons <- list(
  c("Gabon Adults", "Gabon Children"),
  c("Germany Adults", "Germany Children"),
  c("Vietnam Adults", "Vietnam Children")
)

gabon_location_comparisons <- list(
  c("Lambarene Adults", "Lambarene Children"),
  c("Ikobey Adults", "Ikobey Children")
)

plot_country_age_list <- plot_methanogen_abundance_grouped(
  df = combined_df,
  group_var = "CountryGroup",
  comparisons_list = country_comparisons,
  color_by = "country"
)

# Location plots for Gabon
# Subset Gabon samples if needed
gabon_df <- combined_df %>% filter(CurrCountry == "Gabon")

plot_gabon_age_list <- plot_methanogen_abundance_grouped(
  df = gabon_df,
  group_var = "LocationGroup",
  comparisons_list = gabon_location_comparisons,
  color_by = "location"
)

plot_country_age_list
plot_gabon_age_list

ggsave("smithii_age-group_Gabon.pdf", plot_gabon_age_list[["M. smithii"]], width = 60, height = 70, units = "mm",  dpi = 300)
ggsave("smithii_age-group_Gabon.svg", plot_gabon_age_list[["M. smithii"]], width = 60, height = 70, units = "mm",  dpi = 300, device = svglite::svglite)
```


```{r}

```


## Alpha diversity

```{r}
plot_alpha_diversity_by_country <- function(physeq_obj, country_col = "CurrCountry") {
  # Calculate Observed and Shannon diversity
  alpha_df <- estimate_richness(physeq_obj, measures = c("Observed", "Shannon")) %>%
    rownames_to_column("SampleID") %>%
    left_join(as_tibble(sample_data(physeq_obj), rownames = "SampleID"), by = "SampleID")

  # Ensure consistent factor order for countries
  alpha_df[[country_col]] <- factor(alpha_df[[country_col]], levels = c("Gabon", "Germany", "Vietnam"))

  # Rename for consistency
  colnames(alpha_df)[colnames(alpha_df) == "Shannon"] <- "Shannon index"

  # Define comparisons
  comparisons <- list(
    c("Gabon", "Germany"),
    c("Gabon", "Vietnam"),
    c("Germany", "Vietnam")
  )

  # Colors
  country_colors <- c("Gabon" = "#ff0000", "Germany" = "#e7e70e", "Vietnam" = "#0b0bd3")

  # Function to generate plot for one metric
  make_alpha_plot <- function(metric_name, y_label) {
    df_metric <- alpha_df %>%
      select(SampleID, all_of(country_col), value = all_of(metric_name))

    y_max <- max(df_metric$value, na.rm = TRUE)
    y_positions <- seq(y_max * 1.05, length.out = length(comparisons), by = y_max * 0.05)

    ggplot(df_metric, aes_string(x = country_col, y = "value")) +
      geom_boxplot(outlier.shape = NA, fill = NA, color = "black") +
      geom_jitter(aes_string(color = country_col), width = 0.2, alpha = 0.5, size = 1) +
      stat_compare_means(
        comparisons = comparisons,
        method = "wilcox.test",
        label = "p.signif",
        tip.length = 0,
        vjust = 0.3,
        y.position = y_positions
      ) +
      scale_color_manual(values = country_colors) +
      labs(x = NULL, y = y_label) +
      theme_light(base_size = 7, base_family = "Helvetica") +
      theme(
        legend.position = "none",
        axis.text = element_text(size = 7, color = "black"),
        axis.text.x = element_text(size = 7, color = "black"), 
        axis.title = element_text(size = 7, color = "black"),
        plot.title = element_blank()
      )
  }

  # Return both plots in a list
  list(
    Observed = make_alpha_plot("Observed", "Observed richness"),
    Shannon = make_alpha_plot("Shannon index", "Shannon diversity index")
  )
}

```

# Manuscript Figure (Alpha diversity: Fig.1)

```{r}
alpha_plots_adults_all <- plot_alpha_diversity_by_country(ps_adults_raw)
alpha_plots_children_all <- plot_alpha_diversity_by_country(ps_children_raw)

alpha_plots_adults_all$Observed
alpha_plots_adults_all$Shannon
alpha_plots_children_all$Observed
alpha_plots_children_all$Shannon

ggsave("Shannon-index-children.pdf", alpha_plots_children_all$Shannon, width = 95, height = 70, units = "mm", dpi = 300)
ggsave("Shannon-index-children.svg", alpha_plots_children_all$Shannon, width = 95, height = 70, units = "mm", device = svglite::svglite, dpi = 300)
```



```{r}

```

# Manuscript Figure (assessing mutual exclusion M. smithii vs M. intestini: Fig.2A)

```{r}
showtext_auto()
plot_methanogen_scatter <- function(df, abundance_threshold = 0, pseudocount = 1e-5) {

  species_colors <- c(
    "M. smithii" = "#EFC000FF",
    "M. intestini" = "#4A6990FF"
  )

  # Filter and add pseudocounts
  df_filtered <- df %>%
    filter(Abundance > abundance_threshold)

  abundance_wide <- df_filtered %>%
    select(Sample, Species, Abundance, CurrCountry) %>%
    pivot_wider(names_from = Species, values_from = Abundance, values_fill = 0) %>%
    rename(M_smithii = `M. smithii`, M_intestini = `M. intestini`) %>%
    mutate(
      M_smithii = M_smithii + pseudocount,
      M_intestini = M_intestini + pseudocount,
      DominantSpecies = ifelse(M_smithii >= M_intestini, "M. smithii", "M. intestini")
    )

  # Compute Spearman and Kendall correlation
  cor_spearman <- cor.test(abundance_wide$M_smithii, abundance_wide$M_intestini, method = "spearman")
  cor_kendall  <- cor.test(abundance_wide$M_smithii, abundance_wide$M_intestini, method = "kendall")

  best_test <- ifelse(cor_spearman$p.value < cor_kendall$p.value, "Spearman", "Kendall")
  best_rho  <- ifelse(cor_spearman$p.value < cor_kendall$p.value, cor_spearman$estimate, cor_kendall$estimate)
  best_p    <- ifelse(cor_spearman$p.value < cor_kendall$p.value, cor_spearman$p.value, cor_kendall$p.value)

  label_text <- paste0(best_test, " rho = ", round(best_rho, 3), "\n p = ", format(best_p, scientific = TRUE))

  # Plot
  p <- ggplot(abundance_wide, aes(x = M_smithii, y = M_intestini)) +
    geom_point(aes(color = DominantSpecies, shape = CurrCountry), size = 2, alpha = 0.5) +
    geom_smooth(method = "lm", se = TRUE, color = "black", fill = "lightgray") +
    scale_x_log10(labels = scales::label_scientific(digits = 1), name = expression(italic("M. smithii") ~ "abundance")) +
scale_y_log10(labels = scales::label_scientific(digits = 1), name = expression(italic("M. intestini") ~ "abundance")) +
    scale_color_manual(values = species_colors) +
    theme_light(base_size = 7, base_family = "Helvetica") +
    theme(
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
      text = element_text(size = 7, color = "black"),
      legend.position = "top",
      legend.key.size = unit(0.1, "cm"),
      legend.title = element_blank(),
      plot.title = element_text(size = 7, face = "bold")
    ) +
    annotate("text",
             x = max(abundance_wide$M_smithii, na.rm = TRUE) * 0.2,
             y = max(abundance_wide$M_intestini, na.rm = TRUE) * 0.75,
             label = label_text,
             hjust = 0,
             size = 2,
             family = "Helvetica",
             color = "black")

  return(p)
}

plot_scatter_adults <- plot_methanogen_scatter(adults_df, abundance_threshold = 1e-6)
plot_scatter_children <- plot_methanogen_scatter(children_df, abundance_threshold = 1e-6)
plot_scatter_adults
plot_scatter_children

ggsave("positive-cooccurence-children.pdf", plot_scatter_children, width = 90, height = 70, units = "mm", dpi = 300)
ggsave("positive-cooccurence-childrent.svg", plot_scatter_children, width = 90, height = 70, units = "mm", device = svglite::svglite)
```



```{r}

```

# correlation of alpha diversity and methanogen abundance


```{r}
library(lme4)
library(lmerTest)

extract_alpha_methanogen <- function(ps, methanogen_species) {
  # Compute alpha diversity (Shannon)
  alpha_df <- estimate_richness(ps, measures = "Shannon") %>%
    tibble::rownames_to_column("SampleID")
  
  # Extract abundance and taxonomy
  abund_table <- as.data.frame(otu_table(ps))
  tax_table_df <- as.data.frame(tax_table(ps))

  # Match methanogen species
  taxa_to_keep <- rownames(tax_table_df)[
    grepl(methanogen_species, tax_table_df$Species, ignore.case = TRUE)
  ]

  if (length(taxa_to_keep) == 0) {
    stop(paste("No taxa found matching species name:", methanogen_species))
  }

  # Aggregate abundance
  if (taxa_are_rows(ps)) {
    meth_abund <- colSums(abund_table[taxa_to_keep, , drop = FALSE])
  } else {
    meth_abund <- rowSums(abund_table[, taxa_to_keep, drop = FALSE])
  }

  meth_df <- data.frame(
    SampleID = names(meth_abund),
    MethanogenAbundance = meth_abund
  ) %>%
    mutate(ScaledMethanogenAbundance = scale(MethanogenAbundance)[, 1])
  
  # Metadata
  meta_df <- as(sample_data(ps), "data.frame") %>%
    tibble::rownames_to_column("SampleID")

  # Handle AgeYears conversion safely
  if ("Age_years" %in% colnames(meta_df)) {
    meta_df <- meta_df %>%
      mutate(AgeYears = as.numeric(gsub(",", ".", Age_years)))
  } else if ("Age_months" %in% colnames(meta_df)) {
    meta_df <- meta_df %>%
      mutate(AgeYears = as.numeric(gsub(",", ".", Age_months)) / 12)
  } else {
    meta_df$AgeYears <- NA_real_
  }

  # Drop rows with missing age
  meta_df <- meta_df %>%
    filter(!is.na(AgeYears))

  # Merge all
  merged_df <- alpha_df %>%
    inner_join(meth_df, by = "SampleID") %>%
    inner_join(meta_df, by = "SampleID")

  return(merged_df)
}

```

```{r}
# Define the function
fit_alpha_lmm <- function(data, response = "Shannon", predictor = "MethanogenAbundance",
                          fixed_effects = c("CurrCountry"), random_effect = "samplingLocation") {
  
  # Build formula string
  fixed <- paste(c(predictor, fixed_effects), collapse = " + ")
  formula_str <- paste0(response, " ~ ", fixed, " + (1 | ", random_effect, ")")
  model_formula <- as.formula(formula_str)
  
  # Fit linear mixed-effects model
  lmm <- lmerTest::lmer(model_formula, data = data)
  return(summary(lmm))
}
```

```{r}
# For M. smithii in adults
df_adult_smithii <- extract_alpha_methanogen(ps_adults_raw, "Methanobrevibacter_A smithii")
lmm_adult_MS <- fit_alpha_lmm(df_adult_smithii, predictor = "ScaledMethanogenAbundance")
summary(lmm_adult_MS)

# For M. intestini in adults
df_adult_intestini <- extract_alpha_methanogen(ps_adults_raw, "Methanobrevibacter_A smithii_A")
lmm_adult_MI <- fit_alpha_lmm(df_adult_intestini, predictor = "ScaledMethanogenAbundance")
summary(lmm_adult_MI)

# Repeat for children
df_child_smithii <- extract_alpha_methanogen(ps_children_raw, "Methanobrevibacter_A smithii")
lmm_children_MS <- fit_alpha_lmm(df_child_smithii, predictor = "ScaledMethanogenAbundance")
summary(lmm_children_MS)

df_child_intestini <- extract_alpha_methanogen(ps_children_raw, "Methanobrevibacter_A smithii_A")
lmm_children_MI <- fit_alpha_lmm(df_child_intestini, predictor = "ScaledMethanogenAbundance")
summary(lmm_children_MI)

```


