---
title: "O2_SpeicEasi_network_bacteria_archaea"
author: "Mirabeau Mbong"
date: "2025-06-20"
output:
  html_notebook:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
---

This script is used to generate tables used to generate plots in Fig.2B of the associated paper

```{r, setup, include=FALSE}
knitr::opts_chunk$set(dev = "pdf",
                      dpi = 300,
                      echo = FALSE,
                      cache = TRUE)
```

# load libraries

```{r}
library(dplyr)
library(phyloseq)
library(tidyr)

library(SpiecEasi)
threads = 12
seed = 38941

nlambda = 20
lambda.min.ratio = 1e-3
se_reps = 100
```


```{r}
# Check working directory (should be repo root)
message("Current working directory: ", getwd())
```

# Functions

```{r}
create_phyloseq_object <- function(bracken_file, metadata_file) {
  bracken <- read.table(bracken_file, header = TRUE, sep = "\t")
  meta <- read.table(metadata_file, header = TRUE, sep = "\t")

  # Process OTU table
  OTU <- bracken %>%
    dplyr::select(-c(taxonomy_id, taxonomy_lvl, taxonomy, taxIDs)) %>%
    dplyr::select(-contains('frac')) %>%
    dplyr::rename_with(~ sub("[_]num$", "", .x), ends_with("_num"))

  rownames(OTU) <- OTU$name
  OTU <- OTU %>% dplyr::select(-name)

  # Process taxonomy
  tax <- bracken %>%
    dplyr::select(name, taxonomy) %>%
    tidyr::separate(taxonomy,
                    into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
                    sep = ";", fill = "right", extra = "drop")
  rownames(tax) <- tax$name
  tax <- tax %>% dplyr::select(-name)

  # Process metadata
  meta$sample <- meta$ParticipantID
  rownames(meta) <- meta$sample
  meta <- meta %>% dplyr::select(-sample)

  # Create phyloseq object
  ps <- phyloseq(
    otu_table(as.matrix(OTU), taxa_are_rows = TRUE),
    tax_table(as.matrix(tax)),
    sample_data(meta)
  )
  return(ps)
}

```

# prefilter to 20% prevalence across populations

```{r}
filter_phyloseq <- function(ps, min_prevalence = 0.2) {
  library(phyloseq)

  # Subset only the two methanogen species
  ps_methanogen <- subset_taxa(ps, 
    Species %in% c("s__Methanobrevibacter_A smithii", "s__Methanobrevibacter_A smithii_A"))

  # Subset bacteria only
  ps_bacteria <- subset_taxa(ps, Kingdom == "d__Bacteria")

  # Apply prevalence filtering to bacteria only
  ps_bacteria_filtered <- filter_taxa(ps_bacteria, function(x) sum(x > 0) > (min_prevalence * length(x)), TRUE)

  # Merge bacteria (filtered) and methanogens (always kept)
  ps_combined <- merge_phyloseq(ps_bacteria_filtered, ps_methanogen)

  return(ps_combined)
}
```

# run SpiecEasi

```{r}
run_spiec_easi <- function(ps, threads = 12, seed = 38941) {
  library(SpiecEasi)

  spiec.easi(ps, 
             method = 'mb', 
             lambda.min.ratio = 1e-3, 
             nlambda = 20,
             sel.criterion = 'bstars', 
             pulsar.select = TRUE, 
             pulsar.params = list(rep.num = 100, seed = seed, ncores = threads))
}
```

# Format Network with Taxonomy

```{r}
annotate_network <- function(spiec_out, ps) {
  library(igraph)
  library(tidygraph)
  library(stringr)

  # Get edge weights and signs
  optbeta <- as.matrix(symBeta(getOptBeta(spiec_out)))
  weights <- optbeta[upper.tri(optbeta) & optbeta != 0]
  edge_cols <- ifelse(weights > 0, 'positive', 'negative')

  # Get taxonomy table
  tax <- as(tax_table(ps), "matrix") %>% as.data.frame()
  tax$Species <- rownames(tax)  # rownames are species names
  tax <- tax %>%
    mutate(
      name = Species,  # to match graph nodes
      species_2 = str_replace(Species, "s__Methanobrevibacter_A smithii_A", "Methanobrevibacter_A intestini"),
      species_2 = str_replace(species_2, "s__Methanobrevibacter_A smithii", "Methanobrevibacter_A smithii"),
      across(everything(), ~str_remove(.x, "^[a-z]__"))
    )

  # Create graph with species names as node names
  species_names <- tax$name
  Gl <- adj2igraph(getRefit(spiec_out),  
                   rmEmptyNodes = FALSE,
                   vertex.attr = list(name = species_names),
                   edge.attr = list(color = edge_cols, W = weights)) %>%
    as_tbl_graph()

  # Join taxonomy onto graph
  Gl %>%
    activate(nodes) %>%
    inner_join(tax, by = "name") %>%
    mutate(taxonomy = paste(Phylum, Class, Order, sep = ";"))
}

```

# Plot and Save Graph

```{r}
plot_spiec_graph <- function(graph, output_file = NULL) {
  library(ggraph)
  library(ggplot2)

  plot <- graph %>%
    activate(edges) %>%
    mutate(
      abs_W = abs(W),
      edge_linetype = ifelse(W < 0, "dotted", "solid"),
      color = ifelse(W < 0, "negative", "positive")
    ) %>%
    ggraph(layout = 'nicely') + 
    geom_edge_link(aes(color = color, edge_width = abs_W, linetype = color), alpha = 0.3) +
    geom_node_point(aes(color = Class, shape = Kingdom), size = 4) +
    geom_node_text(aes(filter = Kingdom == "Archaea", label = species_2), repel = TRUE) +
    scale_edge_color_manual(name = "Associations", 
                            values = c("negative" = "#969696", "positive" = "forestgreen")) +
    scale_edge_linetype_manual(name = "Associations", 
                               values = c("negative" = "dotted", "positive" = "solid")) +
    scale_edge_width_continuous(name = "Association strength", range = c(0.4, 2)) +
    scale_shape_manual(values = c(15, 16, 17)) +
    theme_graph() +
    theme(
      text = element_text(size = 14),
      legend.box = "horizontal",
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 13)
    )
  
  if (!is.null(output_file)) {
    ggsave(output_file, plot = plot, width = 16, height = 8)
  }
  
  return(plot)
}

```

# Generating Graph Objets (Adult dataset)

```{r}
# Load data
phy_adult <- create_phyloseq_object("data/bracken_output_adults.tsv", "data/filtered_adult_metadata.txt")

# Filter taxa
filtered_adult <- filter_phyloseq(phy_adult)

# Run SpiecEasi
spiec_adult <- run_spiec_easi(filtered_adult)

# Save SpiecEasi result to file
saveRDS(spiec_adult, file = "spiec_adult.rds")

# Annotate graph
graph_adult <- annotate_network(spiec_adult, filtered_adult)

```

# Generating Graph Objets (children dataset)

```{r}
# Load data
phy_children <- create_phyloseq_object("data/bracken_output_children.tsv", "data/filtered_child_metadata.txt")

# Filter taxa
filtered_children <- filter_phyloseq(phy_children)

# Run SpiecEasi
spiec_children <- run_spiec_easi(filtered_children)
# Save SpiecEasi result to file
saveRDS(spiec_children, file = "spiec_children.rds")

# Annotate graph
graph_children <- annotate_network(spiec_children, filtered_children)

```


# Zoom in on methanogen-specific interactions M. smithii or M. intestini with bacteria


```{r}
# Filter Network for Methanogen-Related Edges
filter_methanogen_edges <- function(graph_obj, methanogen_names) {
  edge_list <- graph_obj %>%
    activate(edges) %>%
    as_tibble()

  edge_list_filtered <- edge_list %>%
    filter(from %in% methanogen_names | to %in% methanogen_names)

  node_list <- graph_obj %>%
    activate(nodes) %>%
    as_tibble()

  g_filtered <- tbl_graph(nodes = node_list, edges = edge_list_filtered, directed = TRUE) %>%
    filter(degree(.) > 0)
  
  return(g_filtered)
}

# Annotate Graph with Taxonomy Metadata
annotate_graph_taxonomy <- function(graph_tidy, tax_table_df) {
  tax_table_df <- tax_table_df %>%
    mutate(
      name = rownames(.),  # match graph node names
      label = Species,
      species_2 = ifelse(Species == "s__Methanobrevibacter_A smithii_A", 
                         "s__Methanobrevibacter_A intestini", Species),
      across(c(Species, species_2, Phylum, Kingdom, Class, Family), ~ gsub(".*__", "", .x))
    )

  graph_tidy %>%
    activate(nodes) %>%
    inner_join(tax_table_df, by = "name") %>%
    mutate(taxonomy = paste(Phylum, Class, Order, sep = ";"))
}

#Plot Filtered SpiecEasi Network

plot_spiec_graph_nolegend <- function(graph_annotated, family_colors) {
  graph_annotated <- graph_annotated %>%
    activate(edges) %>%
    mutate(
      abs_W = abs(W),
      edge_category = ifelse(W < 0, "negative", "positive")
    ) %>%
    activate(nodes) %>%
    mutate(
      species_italic = ifelse(
        Kingdom == "Archaea",
        paste0("italic('", species_2, "')"),
        NA
      )
    )

  ggraph(graph_annotated, layout = "nicely") +
    geom_edge_link(
      aes(edge_color = edge_category, edge_width = abs_W),
      edge_alpha = 0.4,
      show.legend = TRUE
    ) +
    geom_node_point(aes(color = Family, shape = Kingdom), size = 3) +
    geom_node_text(
      aes(filter = Kingdom == "Archaea", label = species_italic),
      size = 7 / .pt,  # Convert 7 pt to ggplot size units
      parse = TRUE
    ) +
    scale_shape_manual(
      name = "Kingdom",
      values = c(Archaea = 17, Bacteria = 16)
    ) +
    scale_edge_color_manual(
      name = "Association",
      values = c(positive = "blue", negative = "orchid1")
    ) +
    scale_edge_width_continuous(
      name = "Association strength",
      range = c(0.4, 2),
      breaks = c(0.05, 0.1, 0.2, 0.3),
      labels = scales::number_format(accuracy = 0.1)
    ) +
    scale_color_manual(
      name = "Family",
      values = family_colors
    ) +
    theme_graph(base_family = "Helvetica", base_size = 7) +
    theme(
      legend.position = "none",
      text = element_text(size = 7, family = "Helvetica")
    )
}

```

# Generate M. smithii & M. intestini-bacteria network graphs

```{r}

# IDs of M. smithii and M. intestini in tax_table(filtered_adult)
methanogen_ids <- c(2386, 2387)  

# Create and annotate filtered graph
graph_filtered_adult <- filter_methanogen_edges(graph_adult, methanogen_ids)

# IDs of M. smithii and M. intestini in tax_table(filtered_children)
methanogen_ids1 <- c(2415, 2416)  

# Create and annotate filtered graph
graph_filtered_children <- filter_methanogen_edges(graph_children, methanogen_ids1)

```



```{r}
assign_family_colors <- function(graph_list, color_palette) {
  # Extract all unique families from all graphs
  all_families <- graph_list %>%
    purrr::map(~ as_tibble(activate(.x, "nodes"))$Family) %>%
    unlist() %>%
    unique() %>%
    sort(na.last = TRUE)
  
  n_families <- length(all_families)
  n_colors <- length(color_palette)
  
  if (n_families > n_colors) {
    warning(paste("Number of families (", n_families, ") exceeds number of colors (", n_colors, "). Some colors will be recycled."))
  }
  
  family_colors <- setNames(
    rep(color_palette, length.out = n_families),
    all_families
  )
  
  return(family_colors)
}

family_colors_vec <- c("#E41A1C", "#1E90FF", "#4DAF4A", "#984EA3", "#FFB90F", "#8B636C", "#F781BF", "#999999", 
                       "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#A6D854", "#FFD92F", "#B3B3B3", "#1B9E77", 
                       "#D95F02", "#BF3EFF", "#66CD00", "#E7298A", "#8B814C",  "#A1CAF1", "#A6761D","#BCEE68",
                       "#FF6F61", "#6B5B95", "#88B04B", "#F7CAC9", "#92A8D1", "#955251", "#B565A7", "#97FFFF", 
                       "#104E8B", "#D55E00", "#8B0000")

graphs <- list(graph_filtered_adult, graph_filtered_children)

family_colors <- assign_family_colors(graphs, family_colors_vec)
```


```{r, include=TRUE, fig.width=6, fig.height=4, dpi=300, dev = "pdf"}
# Plot
plot_adult <- plot_spiec_graph_nolegend(graph_filtered_adult, family_colors)

plot_children <- plot_spiec_graph_nolegend(graph_filtered_children, family_colors)

plot_adult
plot_children

ggsave("SpiecEasi_adults_filtered3.pdf", plot_adult, width = 95, height = 75, units = "mm", dpi = 300)
ggsave("SpiecEasi_adults_filtered3.svg", plot_adult, width = 95, height = 75, units = "mm", dpi = 300, device = svglite::svglite)

```

## Generate legends

```{r}
generate_full_family_legend <- function(graph_list, family_colors_vec) {
  # Get unique families from all graphs and assign consistent colors
  family_colors <- assign_family_colors(graph_list, family_colors_vec)

  # Create a dummy data frame for the legend
  family_df <- data.frame(Family = names(family_colors))

  # Generate dummy ggplot for extracting legend
  legend_plot <- ggplot(family_df, aes(x = Family, y = 1, color = Family)) +
    geom_point(size = 3) +
    scale_color_manual(
      name = "Family",
      values = family_colors,
      guide = guide_legend(ncol = 3)
    ) +
    theme_void(base_size = 7, base_family = "Helvetica") +
    theme(
      legend.position = "right",
      legend.text = element_text(family = "Helvetica"),
      legend.title = element_text(family = "Helvetica"),
      legend.spacing.x = unit(0.1, "mm"),
      legend.box.spacing = unit(0.1, "mm")
    )

  # Extract just the legend
  legend_family <- get_legend_only(legend_plot)

  return(list(legend = legend_family, family_colors = family_colors))
}

```


```{r}
library(patchwork)
library(cowplot)

get_legend_only <- function(p) {
  cowplot::get_legend(p)
}
# Combine graphs
graphs <- list(graph_filtered_adult, graph_filtered_children)

# Generate legend and consistent family_colors
legend_out <- generate_full_family_legend(graphs, family_colors_vec)

# Extract outputs
legend_family <- legend_out$legend
family_colors <- legend_out$family_colors  
print(legend_family)

ggsave("Networts_family_legend.pdf", legend_family, width = 90, height = 80, units = "mm", dpi = 300)
ggsave("Networts_family_legend.svg", legend_family, width = 90, height = 80, units = "mm", dpi = 300, device = svglite::svglite)

```

# Extract other figure legends

```{r}
# kingdom legend
legend_kingdom <- ggraph(graph_filtered_adult, layout = "nicely") +
  geom_node_point(aes(shape = Kingdom), size = 4) +
  scale_shape_manual(name = "Kingdom", values = c("Archaea" = 17, "Bacteria" = 16)) +
  theme_void(base_size = 7, base_family = "Helvetica") +
  theme(legend.position = "right")

legend_kingdom <- get_legend_only(legend_kingdom)

# association_legend
 graph_adults <- graph_filtered_adult %>%
  activate(edges) %>%
  mutate(edge_category = ifelse(W < 0, "negative", "positive"))

legend_assoc <- ggraph(graph_adults, layout = "nicely") +
  geom_edge_link(aes(edge_color = edge_category), edge_width = 1, alpha = 0.3) +
  scale_edge_color_manual(name = "Association", values = c("positive" = "blue", "negative" = "orchid1")) +
  theme_void(base_size = 7, base_family = "Helvetica") +
  theme(legend.position = "right")

legend_assoc <- get_legend_only(legend_assoc)

# Association strength legend (edge width)

  graph_adults1 <- graph_filtered_adult %>%
  activate(edges) %>%
  mutate(abs_W = abs(W))

legend_strength <- ggraph(graph_adults1, layout = "nicely") +
  geom_edge_link(aes(edge_width = abs_W), alpha = 0.3) +
  scale_edge_width_continuous(
    name = "Association strength",
    range = c(0.4, 2),
    breaks = c(0.05, 0.1, 0.2, 0.3, 0.4, 0.5),
    labels = scales::number_format(accuracy = 0.1)
  ) +
  theme_void(base_size = 7, base_family = "Helvetica") +
  theme(legend.position = "right")

legend_strength <- get_legend_only(legend_strength)

ggsave("legend_associations_strength_adults_network.pdf", legend_strength, width = 30, height = 45, units = "mm", dpi = 300)
ggsave("legend_associations_strength_adults_network.svg", legend_strength, width = 30, height = 45, units = "mm", dpi = 300, device = svglite::svglite)



```

# Manuscript Figure (Extract Edge list Fig.2)

```{r}
extract_edge_list <- function(graph_obj, physeq_obj) {
  # Extract edge list
  edge_list <- graph_obj %>%
    activate(edges) %>%
    as_tibble()
  
  # Extract node metadata
  node_data <- graph_obj %>%
    activate(nodes) %>%
    as_tibble()
  
  # Attach node names to edge list
  edge_list_named <- edge_list %>%
    mutate(
      from_name = node_data$name[from],
      to_name = node_data$name[to]
    ) %>%
    rename(
      association_strength = W,
      association = color
    ) %>%
    select(
      bacteria_species = from_name,
      methanogen_species = to_name,
      association_strength,
      association
    ) %>%
    mutate(Species = bacteria_species)

  # Extract taxonomy table
  taxonomy_df <- as.data.frame(tax_table(physeq_obj))

  # Clean taxonomic labels: strip prefixes (e.g., "s__") in Species column
  taxonomy_df <- taxonomy_df %>%
    as.data.frame() %>%
    rownames_to_column(var = "ASV") %>%
    mutate(Species = gsub("^[a-z]__+", "", Species))  # Remove prefixes like s__, g__, etc

  # Add ASV ID to match with edge list species 
  taxonomy_df <- taxonomy_df %>%
    mutate(node_name = gsub("^[a-z]__+", "", Species))

  # Join taxonomy info using cleaned names
  final_edge_list <- edge_list_named %>%
    left_join(taxonomy_df, by = c("Species" = "node_name"))

  return(final_edge_list)
}

```


# Write Edge list to tables (Manuscript table, also used to generate Figure 2B)

```{r}
# For children
edge_table_children <- extract_edge_list(graph_filtered_children, filtered_children)

# For adults
edge_table_adults <- extract_edge_list(graph_filtered_adult, filtered_adult)

# Save as TSV or Excel 
readr::write_tsv(edge_table_children, "Supplementary_Co-ocuurence-edge_table_children.tsv")
readr::write_tsv(edge_table_adults, "Supplementary_Co-ocuurence-edge_table_Adults.tsv")
```

