---
title: "03_Genome_stats_cultured_strains"
author: "Mirabeau Mbong"
date: "2025-06-27"
output:
  html_notebook:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
---

This script is used to generate plots in Fig.3B & C and Fig.4 of the associated paper

```{r, setup, include=FALSE}
knitr::opts_chunk$set(dev = "pdf",
                      dpi = 300,
                      echo = FALSE,
                      cache = TRUE)

```

# Load libraries

```{r}
library(tidyverse)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
library(svglite)
```


```{r}
# Check working directory (should be repo root)
message("Current working directory: ", getwd())
```


```{r}

# Read in your main metadata and CheckM2 quality report
meta <- read_tsv("data/metadata_all_available_genomes.txt")
checkm <- read_tsv("data/CheckM2_quality_report.tsv")

# Merge both by genome ID (Name column is common)
merged <- left_join(meta, checkm, by = "Name") #%>%

```

# Add M. smithii type strain metadata (ATCC 35061)

```{r}
merged <- merged %>%
  mutate(corrected_genome_size = (Genome_Size * 100 / (Completeness)) - (Genome_Size * (Contamination / 100)))

ref <- read_tsv("data/drep_genomes_metadata_2.txt")

ref_atcc <- ref %>%
  filter(country == "NCBI ref") %>%
  mutate(country = "ATCC 35061",
         Name = isolate,
         genomeID = isolate,  # set genomeID to match Name/isolate
         # Using correct completeness and contamination values
         corrected_genome_size = (genome.size * 100 / 97.71) - (genome.size * 2.51 / 100)) %>%
  select(Name, genomeID, genome.size, total_genes, corrected_genome_size, country, species) %>%
  rename(Genome_Size = genome.size, Total_Coding_Sequences = total_genes)

# Add to merged table
merged_full <- bind_rows(merged, ref_atcc)
colnames(merged_full)

```

# Manuscript figaure (Completion vs Contamination: Fig.3B)

```{r}
# Set factor levels in the order you want the legend
merged$species <- factor(merged$species, levels = c("M. smithii", "M. intestini", "M. oralis"))

# Custom colors, must be named with matching factor levels
species_colors1 <- c(
  "M. smithii" = "#EFC000FF",
  "M. intestini" = "#4A6990FF",
  "M. oralis" = "blue"
)

# Legend labels as expressions (note: do NOT quote them, and do NOT name them)
species_labels_italic <- expression(italic("M. smithii"), italic("M. intestini"), italic("M. oralis"))

# Plot
quality_plot <- ggplot(merged, aes(x = Completeness, y = Contamination, color = species)) +
  geom_point(size = 2, alpha = 0.5) +
  scale_color_manual(
    name = "Species",
    values = species_colors1,
    labels = species_labels_italic
  ) +
  theme_light(base_size = 7, base_family = "Helvetica") +
  labs(x = "Completeness (%)", y = "Contamination (%)", title = "") +
  theme(
    axis.text = element_text(size = 7),
    axis.title = element_text(size = 7),
    strip.text = element_text(size = 7),
    plot.title = element_text(size = 7, colour = "black"),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7),
    legend.text.align = 0)

quality_plot
ggsave("CheckM2_genome_quality-all-strains.pdf", quality_plot, width = 65, height = 55, units = "mm", dpi = 300)
ggsave("CheckM2_genome_quality-all-strains.svg", quality_plot, width = 65, height = 55, units = "mm", dpi = 300, device = svglite::svglite)
```

```{r}

```


# Manuscript figure (Species Isolation by Country: Fig.3C)

```{r}
species_bar_plot <- merged %>%
  filter(!is.na(country)) %>%
  count(country, species) %>%
  ggplot(aes(x = country, y = n, fill = species)) +
  geom_bar(stat = "identity", width = 0.3) +
  scale_fill_manual(
    name = "Species",
    values = species_colors1,
    labels = species_labels_italic
  ) +
  coord_flip() +  # Flip axes
  theme_light(base_size = 7, base_family = "Helvetica") +
  labs(
    y = "Number of isolates",
    x = "",
    title = ""
  ) +
  theme(
    axis.text = element_text(size = 7),
    axis.title = element_text(size = 7),
    strip.text = element_text(size = 7),
    plot.title = element_text(size = 7, colour = "black"),
    legend.text = element_text(size = 7),
    legend.title = element_blank(),
    legend.position = "top",
    legend.key.size = unit(0.3, "cm"),
    panel.grid.major.y = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

# Display the plot
print(species_bar_plot)

# Save it to file
ggsave("species_isolation_by_country.pdf", species_bar_plot, width = 65, height = 55, units = "mm", device = cairo_pdf)
ggsave("species_isolation_by_country.svg", species_bar_plot, width = 65, height = 55, units = "mm", dpi = 300, device = svglite::svglite)


```


```{r}

```

# gene counts, average gene length, corrected genome size, GC content, coding density


```{r}
plot_genome_comparisons <- function(df, 
                                      metric_col, 
                                      y_label = NULL, 
                                      save_path = NULL, 
                                      width = 4, 
                                      height = 4, 
                                      dpi = 300) {
  # Filter relevant species
  df_filtered <- df %>%
    filter(species %in% c("M. smithii", "M. intestini")) %>%
    mutate(
      species = factor(species, levels = c("M. smithii", "M. intestini")),
      Genome_Size_Mbp = corrected_genome_size / 1e6
    )

  # Validate metric
  if (!(metric_col %in% colnames(df_filtered))) {
    stop(paste("Metric", metric_col, "not found in dataframe."))
  }

  # Color palette
  species_colors <- c(
    "M. smithii" = "#EFC000FF",
    "M. intestini" = "#4A6990FF"
  )

  # Plot
  p <- ggplot(df_filtered, aes(x = species, y = .data[[metric_col]])) +
    geom_boxplot(outlier.shape = NA, fill = "white", color = "black", alpha = 0.5) +
    geom_jitter(aes(color = species), width = 0.15, alpha = 0.5, size = 1.5) +
    stat_summary(
      fun = median,
      geom = "text",
      aes(label = round(after_stat(y), 2)),
      position = position_nudge(x = 0.15),
      vjust = -0.3,
      hjust = -0.05,
      size = 2,
      fontface = "bold",
      color = "blue"
    ) +
    scale_color_manual(values = species_colors) +
    theme_light(base_size = 7, base_family = "Helvetica") +
    scale_x_discrete(
      labels = parse(text = c("italic('M. smithii')", "italic('M. intestini')"))
    ) +
    labs(x = NULL, y = y_label %||% metric_col, color = "Species") +
    theme(
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
      strip.text = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 7),
      legend.position = "none",
      panel.grid.major.y = element_line(color = "gray90"),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )

  # Save if path provided
  if (!is.null(save_path)) {
    ggsave(save_path, plot = p, width = width, height = height, dpi = dpi, bg = "white")
  }

  return(p)
}

```


# Manuscript figaure (cultured strain features: using unique strains:ANI = 99.99; Fig.4)

```{r}
meta_drep <- read_tsv("data/Genome_stats_86-strains.txt")
colnames(meta_drep)
Coding_Density = plot_genome_comparisons(df = meta_drep, 
                          metric_col = "Coding_Density", 
                          y_label = "Coding Density")

# Plot for Total Coding Sequences
Total_Coding_Sequences = plot_genome_comparisons(df = meta_drep, 
                          metric_col = "Total_Coding_Sequences", 
                          y_label = "Total Coding Sequences")

corrected_genome_size = plot_genome_comparisons(df = meta_drep, 
                          metric_col = "Genome_Size_Mbp", 
                          y_label = "Genome size (Mbp)")

Average_Gene_Length = plot_genome_comparisons(df = meta_drep, 
                          metric_col = "Average_Gene_Length", 
                          y_label = "Average Gene Length")

GC_Content = plot_genome_comparisons(df = meta_drep, 
                          metric_col = "GC_Content", 
                          y_label = "GC Content")

Coding_Density
Total_Coding_Sequences
corrected_genome_size
Average_Gene_Length
GC_Content

ggsave("GC_Content.pdf", GC_Content, width = 55, height = 45, units = "mm", dpi = 300)
ggsave("GC_Content.svg", GC_Content, width = 55, height = 45, units = "mm", dpi = 300, device = svglite::svglite)

```



```{r}
plot_genome_size_by_country <- function(df, y_positions = c(2.35, 2.45, 2.55), save_path = NULL, width = 4, height = 4, dpi = 300) {
  # Define country colors
  country_colors <- c("Gabon" = "#ff0000", "Germany" = "#e7e70e", "Vietnam" = "#0b0bd3")
  
  # Define pairwise comparisons
  comparisons <- list(
    c("Gabon", "Germany"),
    c("Gabon", "Vietnam"),
    c("Germany", "Vietnam")
  )
  
  # Prepare data
  df_filtered <- df %>%
    filter(species == "M. smithii") %>%
    mutate(
      Country = case_when(
        grepl("^G", genome) ~ "Gabon",
        grepl("^T", genome) ~ "Germany",
        grepl("^V", genome) ~ "Vietnam",
        TRUE ~ NA_character_
      ),
      Genome_Size_Mbp = corrected_genome_size / 1e6
    ) %>%
    filter(Country %in% c("Gabon", "Germany", "Vietnam"))

  # Plot
  p <- ggplot(df_filtered, aes(x = Country, y = Genome_Size_Mbp)) +
    geom_boxplot(fill = NA, color = "black", outlier.shape = NA) +  # No fill, only outline
    geom_jitter(aes(color = Country), width = 0.15, alpha = 0.5, size = 1.5) +
    stat_summary(
      fun = median,
      geom = "text",
      aes(label = round(after_stat(y), 2)),
      position = position_nudge(x = 0.25),
      vjust = -0.3,
      size = 2,
      fontface = "bold",
      color = "blue"
    ) +
    stat_compare_means(
      comparisons = comparisons,
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0,
      vjust = 0.3,
      y.position = y_positions
    ) +
    scale_color_manual(values = country_colors) +
    theme_light(base_size = 7, base_family = "Helvetica") +
    labs(
      x = NULL,
      y = "Corrected Genome Size (Mbp)"
    ) +
    theme(
      axis.text.x = element_text(size = 7, family = "Helvetica"),
      axis.text.y = element_text(size = 7, family = "Helvetica"),
      axis.text = element_text(size = 7, family = "Helvetica"),
      axis.title = element_text(size = 7, family = "Helvetica"),
      strip.text = element_text(size = 7, family = "Helvetica"),
      legend.position = "none",
      panel.grid.major.y = element_line(color = "gray90"),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
  
  # Save if path is provided
  if (!is.null(save_path)) {
    ggsave(save_path, plot = p, width = width, height = height, dpi = dpi, bg = "white")
  }
  
  return(p)
}

```

# Manuscript figure (M. smithii corrected genome size by country: Fig.4E)

```{r}
genome_size_M_smithii = plot_genome_size_by_country(df = meta_drep, y_positions = c(2.35, 2.45, 2.55))
genome_size_M_smithii
ggsave("Genome_size_m_smithii_by_county.pdf", genome_size_M_smithii, width = 60, height = 50, units = "mm", dpi = 300)
ggsave("Genome_size_m_smithii_by_county.svg", genome_size_M_smithii, width = 60, height = 50, units = "mm", dpi = 300, device = svglite::svglite)
```
```{r}

```

# Manuscript figure (M. intestini corrected genome size by country: Fig.4F)
```{r}
meta_mi <- read_tsv("data/Genome_stats_48_M_intestini.txt")
meta_mi$geo_category[meta_mi$geo_category == "africa"] <- "African"
meta_mi$geo_category[meta_mi$geo_category == "western world"] <- "Eurasian"

plot_genome_size_by_geo_category <- function(
  df,
  y_positions = c(2.45),  # Only one comparison
  save_path = NULL,
  width = 4,
  height = 4,
  dpi = 300
) {
  # Define geo category colors
  geo_colors <- c("African" = "#F8766D", "Eurasian" = "#00BFC4")

  # Define pairwise comparison
  comparisons <- list(c("African", "Eurasian"))

  # Prepare data
  df_filtered <- df %>%
    filter(species == "M. intestini") %>%
    mutate(
      Genome_Size_Mbp = corrected_genome_size / 1e6,
      geo_category = case_when(
        geo_category == "AfricaN" ~ "African",
        geo_category == "Eurasian" ~ "Eurasian",
        TRUE ~ geo_category
      )
    ) %>%
    filter(geo_category %in% c("African", "Eurasian"))

  # Plot
  p <- ggplot(df_filtered, aes(x = geo_category, y = Genome_Size_Mbp)) +
    geom_boxplot(fill = NA, color = "black", outlier.shape = NA) +
    geom_jitter(aes(color = geo_category), width = 0.15, alpha = 0.5, size = 1.5) +
    stat_summary(
      fun = median,
      geom = "text",
      aes(label = round(after_stat(y), 2)),
      position = position_nudge(x = 0.25),
      vjust = -0.3,
      size = 2,
      fontface = "bold",
      color = "blue"
    ) +
    stat_compare_means(
      comparisons = comparisons,
      method = "wilcox.test",
      label = "p.signif",
      tip.length = 0,
      vjust = 0.3,
      y.position = y_positions
    ) +
    scale_color_manual(values = geo_colors) +
    theme_light(base_size = 7, base_family = "Helvetica") +
    labs(
      x = NULL,
      y = "Corrected Genome Size (Mbp)"
    ) +
    theme(
      axis.text = element_text(size = 7, family = "Helvetica"),
      axis.title = element_text(size = 7, family = "Helvetica"),
      strip.text = element_text(size = 7, family = "Helvetica"),
      legend.position = "none",
      panel.grid.major.y = element_line(color = "gray90"),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )

  # Save if path is provided
  if (!is.null(save_path)) {
    ggsave(save_path, plot = p, width = width, height = height, dpi = dpi, bg = "white")
  }

  return(p)
}

genome_size_M_intestini = plot_genome_size_by_geo_category(
  df = meta_mi,
  y_positions = c(2.2)  
)

genome_size_M_intestini
ggsave("Genome_size_m_intestini_by_county.pdf", genome_size_M_intestini, width = 80, height = 65, units = "mm", dpi = 300)
ggsave("Genome_size_m_intestini_by_county.svg", genome_size_M_intestini, width = 80, height = 65, units = "mm", dpi = 300, device = svglite::svglite)
```


